---
title: "Sea Duck Incubation Constancy Analysis: 2011–2025"
subtitle: "Draft Summaries and Analyses - updated Jan"
author: "Kayla Shively, USFWS Collaborators"
format:
  html:
    embed-resources: true
    code-fold: true      
execute:
  warning: false
  message: false
  echo: true
params:
  hen_file:   "data/hen_activities_current.csv"
  nests_file: "data/nests_final_20251201.csv"
  met_file:   "data/barrow_ghcnd_daily.csv"
---

```{r setup}

# Load libraries
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(gt)
library(xfun)
library(rstatix)
library(coin)
library(patchwork)
library(gridExtra)
library(glmmTMB)
library(performance)
library(purrr)
library(broom)
library(glue)
library(car)
library(cowplot)
library(broom.mixed)
library(DHARMa)
library(FSA)
library(tidyr)
library(gtable)
library(scales)
library(dplyr)
library(tibble)
library(ggh4x)
library(lmerTest)
library(MuMIn)
```

```{r}
# Read data from relative paths defined in YAML params
hen.act <- read.csv(params$hen_file)

nests <- read.csv(params$nests_file)

met_raw <- read.csv(params$met_file)

# Set color palette
palette <- c(
  "KIEI" = "#4DBBD5FF",  
  "LTDU" = "#3C5488FF",  
  "SPEI" = "#00A087FF", 
  "STEI" = "#E64B35FF"   
)

pal_common <- c(
  "King Eider"         = unname(palette["KIEI"]),
  "Long-tailed Duck"   = unname(palette["LTDU"]),
  "Steller's Eider"    = unname(palette["STEI"]),
  "Spectacled Eider"   = unname(palette["SPEI"]))

# add common names
species_labels <- c(
  KIEI = "King Eider",
  LTDU = "Long-tailed Duck",
  STEI = "Steller's Eider",
  SPEI = "Spectacled Eider"
)

```

## Introduction

This document summarizes incubation constancy patterns, recess behavior in relation to environmental effects and body-mass relationships across four sea duck species monitored between **2011 and 2025**:

-   **King Eider (KIEI)**

-   **Long-tailed Duck (LTDU)**

-   **Spectacled Eider (SPEI)**

-   **Steller’s Eider (STEI)**

All analyses update as underlying data files are updated.

No data are available for years 2017, 2018, 2020, and 2021. For 2017 and 2018, Graff 2018; 2021 report that 18 KIEI and 8 LTDU nest had nest cameras, but they all had experimental nest covers, which I assume is why they are not included in the Access database for incubation behavior. No surveys were conducted during the Covid19 pandemic in 2020 and 2021.

**Biological Hypotheses Tested**

1.  There are species-specific differences in incubation constancy (e.g., larger-bodied ducks like KIEI and SPEI will have higher incubation metrics than LTDU and STEI).
2.  Weather impact daily incubation constancy.
3.  Incubation constancy is not constant throughout the nesting season.
4.  Incubation constancy differs for Hatched and Failed nests.

## Nest Monitoring Overview

A total of `` `r dplyr::n_distinct(hen.act$Nest_ID)` ``active nests were located across all years, consisting of:

-   `` `r dplyr::n_distinct(hen.act$Nest_ID[hen.act$Species == "KIEI"])` `` **KIEI**

-   `` `r dplyr::n_distinct(hen.act$Nest_ID[hen.act$Species == "LTDU"])` `` **LTDU**

-   `` `r dplyr::n_distinct(hen.act$Nest_ID[hen.act$Species == "SPEI"])` `` **SPEI**

-   `` `r dplyr::n_distinct(hen.act$Nest_ID[hen.act$Species == "STEI"])` `` **STEI**

```{r}
### Filtering

## Add Duration Mins to hen.act
hen.act <- hen.act %>%
  mutate(
    DurationSecs = as.numeric(DurationSecs),
    DurationMins = if_else(
      !is.na(DurationSecs),
      DurationSecs / 60,
      NA_real_
    )
  )

## Remove nests with < 1 day (1440 mins) of observation
nest.incub.time <- hen.act %>%
  filter(Behavior %in% c("N-NotIncubating", "I-Incubating")) %>%
  group_by(Nest_ID) %>%
  summarise(
    total.incub.mins = sum(DurationMins, na.rm = TRUE),
    .groups = "drop"
  )

# Identify nests with >1440 mins
valid.nests <- nest.incub.time %>%
  filter(total.incub.mins >= 1440) %>%
  pull(Nest_ID)

# Identify nests with >1440 mins to view
invalid.nests <- nest.incub.time %>%
  filter(total.incub.mins < 1440) %>%
  distinct(Nest_ID)

# New working dataset without low nests
hen.act.filtered <- hen.act %>%
  filter(Nest_ID %in% valid.nests) %>%
  filter(Nest_ID != "23-DRS009") # TEMP removal until time error resolved
```

**Filtering**

-   `` `r length(unique(hen.act$Nest_ID)) - length(unique(hen.act.filtered$Nest_ID))` `` nests were removed from the dataset because they had \<1 day (1440 minutes) of incubation behavior (failed shortly after deployment).

**Table 1.**  Number of sea duck nests monitored with time-lapse cameras from 2011-2025, after filtering nests with less than 1 day of incubation monitoring. Apparent nest success is calculated as the proportion of nests that hatched among those with known fates. Years without data reflect periods of experimental nest covers or suspended field operations (2017, 2018, 2020, and 2021).

```{r}

# Drop nests with undetermined fates for summary
hen.act.filtered <- hen.act.filtered %>%
  filter(!Fate_combined %in% c(
    "Undetermined - Nest is not viewable",
    "Undetermined - Active upon cam collect or battery died"
  ))

# Triple check no NAs
hen.act.filtered <- hen.act.filtered %>%
  filter(!is.na(Nest_ID), !is.na(Species))

# New column with just hatch/fail
hen.act.filtered <- hen.act.filtered %>%
  mutate(
    Fate_simple = case_when(
      Fate_combined == "Hatched" ~ "Hatched",
      Fate_combined %in% c(
        "Depredated",
        "Abandoned",
        "Trampled - Caribou herd trampling area where nest is"
      ) ~ "Failed",
      TRUE ~ NA_character_
    )
  )

# Create one record per nest. 
nest.level <- hen.act.filtered %>%
  group_by(Species, year, Nest_ID) %>%
  summarise(
    Fate_simple = case_when(
      any(Fate_combined == "Hatched", na.rm = TRUE) ~ "Hatched",
      any(Fate_combined %in% c(
        "Depredated",
        "Abandoned",
        "Trampled - Caribou herd trampling area where nest is"
      ), na.rm = TRUE) ~ "Failed",
      TRUE ~ NA_character_
    ),
    .groups = "drop"
  )

# Count nests per species per year
nests.per.year <- nest.level %>%
  count(Species, year, name = "n_nests")

# Calculate apparent nest success
apparent.success <- nest.level %>%
  filter(!is.na(Fate_simple)) %>%      # only nests with determinable fate
  group_by(Species) %>%
  summarise(
    apparent.success      = mean(Fate_simple == "Hatched"),
    apparent_success_pct  = round(100 * apparent.success, 1),
    .groups = "drop"
  )

# Convert to wide-format
nests.wide <- nests.per.year %>%
  pivot_wider(
    names_from  = year,
    values_from = n_nests,
    values_fill = 0
  )

# Order my columns
year_cols <- nests.wide %>%
  select(-Species) %>%
  names() %>%
  as.integer() %>%
  sort() %>%
  as.character()

nests.wide.ordered <- nests.wide %>%
  select(Species, all_of(year_cols))

# Combine tables
summary_wide <- nests.wide.ordered %>%
  left_join(apparent.success, by = "Species") %>%
  mutate(
    Total = rowSums(across(all_of(year_cols)), na.rm = TRUE)
  ) %>%
  relocate(Total, .after = all_of(tail(year_cols, 1))) %>%
  arrange(Species)

# Display publication-quality table with common names
nest_summary <- summary_wide %>%
  mutate(
    Species = dplyr::recode(
      as.character(Species),
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      STEI = "Steller's Eider",
      SPEI = "Spectacled Eider"
    )
  ) %>%
  select(Species, all_of(year_cols), Total, apparent_success_pct) %>%
  rename(`Apparent Nest Success (%)` = apparent_success_pct) %>%
  gt() %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )

nest_summary

gtsave(
  nest_summary,
  filename = "table_nest_summary_by_year.png",
  vwidth  = 900,
  vheight = 300
)
```

```{r}
##### NOTE: There was just 1 nest with an UNK predator, 22-ESW096, and it was filtered out due to <1440 mins of video time. 


# Summarize by Year and Predator, excluding "NotApplicable"
predator_summary <- hen.act %>%
  filter(!is.na(Fate_Dep_Species_video),
         Fate_Dep_Species_video != "NotApplicable") %>%
  group_by(year, Fate_Dep_Species_video) %>%
  summarise(
    n_nests = n_distinct(Nest_ID),
    .groups = "drop"
  )

# Add total nests per year for labeling purposes
predator_summary <- predator_summary %>%
  group_by(year) %>%
  mutate(total_nests_year = sum(n_nests)) %>%
  ungroup()

# Create lookup table for x-axis labels
year_counts <- predator_summary %>%
  distinct(year, total_nests_year)

# Create "(n = ...)" style labels
year_labels <- setNames(
  paste0(year_counts$year, " (n = ", year_counts$total_nests_year, ")"),
  year_counts$year
)

# Identify numeric x positions for fox control years
years_to_box <- 2012:2016
x_levels <- levels(factor(predator_summary$year))
xmin <- which(x_levels == "2011") - 0.5
xmax <- which(x_levels == "2016") + 0.5

# Plot raw counts as stacked bars with red box highlight for fox dep years
ggplot(predator_summary, aes(x = factor(year), y = n_nests, fill = Fate_Dep_Species_video)) +
  geom_rect(
    inherit.aes = FALSE,
    xmin = xmin, xmax = xmax,
    ymin = -Inf, ymax = Inf,
    fill = NA, color = "red", linewidth = 1
  ) +
  geom_col() +
  labs(
    x = "Year",
    y = "Number of depredated nests",
    fill = "Predator species"
  ) +
  scale_x_discrete(labels = year_labels) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

**Figure. EXCLUDED** Stacked bars show the count of nests predated during years between 2011 - 2025. Colors correspond to predator species detected via camera evidence or field sign, including Arctic Fox (ARFO), Raven (REFO), Glaucous Gull (GLGU), Jaeger species (JASP), Long-tailed Jaeger (LTJA), Parasitic Jaeger (PAJA), Polar Bear (POBE), Pomarine Jaeger (POJA), and Snowy Owl (SNOW). Considerable interannual variation is evidence, with Arctic Fox consistently responsible for a large share of depredations in most years, while other species (e.g., gulls, jaegers, polar bears) contribute episodically. Arctic fox control was carried out by USDA for 12 consecutive years and including 2011-2016. There were no nests with Unknown predators.

```{r}

# Remove replicate rows per individual recess
recess_level <- hen.act.filtered %>%
  filter(Behavior == "N-NotIncubating") %>%
  distinct(Nest_ID, Species, RecessCause, ID, DurationMins)

# Recess-level table for unique recess events
recess_level <- hen.act.filtered %>%
  filter(Behavior == "N-NotIncubating") %>%
  mutate(RecessDate = as.Date(DateTimeBegin)) %>%
  distinct(Nest_ID, Species, RecessCause, ID, RecessDate, DurationMins)

# Mean recess duration per activity, per species
summary_duration <- recess_level %>%
  group_by(Species, RecessCause) %>%
  summarise(
    n_recesses = sum(!is.na(DurationMins)),
    mean_value = mean(DurationMins, na.rm = TRUE),
    sd = sd(DurationMins, na.rm = TRUE),
    se = sd / sqrt(n_recesses),
    t_crit = qt(0.975, df = n_recesses - 1),
    lower_CI = mean_value - t_crit * se,
    upper_CI = mean_value + t_crit * se,
    .groups = "drop"
  ) %>%
  mutate(
    metric = "Mean recess duration (min)",
    lower_CI = pmax(lower_CI, 0)
  )

# Recesses per day, per nest-day then summarized
recesses_per_day <- recess_level %>%
  group_by(Nest_ID, Species, RecessCause, RecessDate) %>%
  summarise(
    recesses_that_day = n(),
    .groups = "drop"
  )

summary_recesses_per_day <- recesses_per_day %>%
  group_by(Species, RecessCause) %>%
  summarise(
    n_days = n(),
    mean_value = mean(recesses_that_day, na.rm = TRUE),
    sd = sd(recesses_that_day, na.rm = TRUE),
    se = sd / sqrt(n_days),
    t_crit = qt(0.975, df = n_days - 1),
    lower_CI = mean_value - t_crit * se,
    upper_CI = mean_value + t_crit * se,
    .groups = "drop"
  ) %>%
  mutate(
    metric = "Recesses per day (count)",
    lower_CI = pmax(lower_CI, 0)
  )

# Combine plots as stacked panels
fig2_table <- bind_rows(summary_duration, summary_recesses_per_day)

ggplot(fig2_table, aes(x = RecessCause, y = mean_value, color = Species)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbar(
    aes(ymin = lower_CI, ymax = upper_CI),
    position = position_dodge(width = 0.6),
    width = 0.15
  ) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = palette) +
  scale_x_discrete(labels = c("H" = "Human", "O" = "Other", "P" = "Predator")) +
  labs(
    title = "",
    x = "Recess Cause",
    y = NULL,
    color = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

**Figure EXCLUDED.** Mean duration of incubation recesses with standard error for four sea duck species across three categories of recess cause: human disturbance (researchers), other causes (e.g., voluntary recesses, undetermined causes), and predator disturbance. For most ducks, predator disturbance tends to lead to longer and more variable recess durations. KIEI and LTDU had particularly long predator-related recesses. KIEI and LTDU were more vulnerable to longer recesses caused by humans. The shorter "Other" recesses were shorter, as expected.

**Table 2.** Coverage of time-lapse video monitoring of incubation behavior in Arctic-breeding sea duck nests near Utqiagvik, Alaska, USA between 2011-2025.

```{r}

# Verify fixed quirk with DurationMins, create days column
hen.act.filtered <- hen.act.filtered %>%
  mutate(
    DateTimeBegin = as.POSIXct(DateTimeBegin, tz = "UTC"),
    DateTimeEnd   = as.POSIXct(DateTimeEnd,   tz = "UTC"),
    DurationMins  = as.numeric(difftime(DateTimeEnd, DateTimeBegin, units = "mins")),
    DurationDays  = DurationMins / (60 * 24)   # convert minutes → days
  )

# Summarize days per nest of I and N behaviors
nest_days <- hen.act.filtered %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(Species, Nest_ID) %>%
  summarise(
    days_per_nest = sum(DurationDays, na.rm = TRUE),
    .groups = "drop"
  )

# Summarize by species and add SD
nest_summary <- nest_days %>%
  group_by(Species) %>%
  summarise(
    Number_of_Nests = n(),
    Total_Days_Monitored = sum(days_per_nest, na.rm = TRUE),
    Mean_Days = mean(days_per_nest, na.rm = TRUE),
    SD_Days = sd(days_per_nest, na.rm = TRUE),
    Mean_SD = paste0(
      round(Mean_Days, 1), " ± ", round(SD_Days, 1)
    ),
    .groups = "drop"
  )

# Create table with mean ± SD instead of separate columns
nest_coverage <- nest_summary %>%
  mutate(
    Species = factor(
      Species,
      levels = names(species_labels),
      labels = unname(species_labels)
    )
  ) %>%
  select(
    Species,
    `N (nests)` = Number_of_Nests,
    `Days Monitored` = Total_Days_Monitored,
    `Mean (± SD)` = Mean_SD
  ) %>%
  gt() %>%
  cols_align(align = "center", columns = everything()) %>%
  fmt_number(columns = c(`N (nests)`, `Days Monitored`), decimals = 0) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )
nest_coverage

gtsave(
  nest_coverage,
  filename = "table_nest_coverage.png",
  vwidth  = 900,
  vheight = 300
)
```

```{r}
# Note on nesting initiation. Completeness varies among incubation initiation estimates (candling, hatch observations, etc), so I've made a tiered priority structure for estimating incubation start: inc_firsthatch; inc_hatch; inc_mean. If hatch is observed, we use inc_firsthatch, else inc_hatch. If candling/floating data are available, we use inc_mean. I created a new column that estimates this.

# Prepping dataframe
# Tiered incubation estimate strategy.
hen.act.filtered <- hen.act.filtered %>%
  mutate(
    inc_init_final = case_when(
      !is.na(inc_firsthatch) ~ inc_firsthatch,
      !is.na(inc_hatch) ~ inc_hatch,
      !is.na(inc_mean) ~ inc_mean,
      TRUE ~ NA_real_
    ),
    inc_init_tier = case_when(
      !is.na(inc_firsthatch) ~ "hatch_first",
      !is.na(inc_hatch) ~ "hatch_mean",
      !is.na(inc_mean) ~ "floating_mean",
      TRUE ~ "missing"
    )
  )

# Calculating Day of Year of each activity. 
hen.act.filtered <- hen.act.filtered %>%
  mutate(
    DateTimeBegin = parse_date_time(DateTimeBegin,
                                    orders = c("Ymd HMS", "Ymd HM", "Ymd", 
                                               "mdY HMS", "mdY HM", "mdY")),
    doy = yday(DateTimeBegin)
  )

# Calculate days since initiation
hen.act.filtered <- hen.act.filtered %>%
  mutate(
    DateTimeBegin = ymd_hms(DateTimeBegin),
    doy           = yday(DateTimeBegin),
    day_since_init = doy - inc_init_final,
    day_since_init_int = floor(day_since_init)
  )


# Build dataset for nests monitored per day since initiation
monitoring_dist <- hen.act.filtered %>%
  filter(
    Behavior %in% c("I-Incubating", "N-NotIncubating"),
    !is.na(day_since_init_int)
  ) %>%
  distinct(Species, Nest_ID, day_since_init_int) %>%  # one record per nest per day
  count(Species, day_since_init_int, name = "n_nests")


plot_dat <- monitoring_dist %>%
  mutate(
    Species = trimws(as.character(Species)),
    Species = case_when(
      Species == "KIEI" ~ "King Eider",
      Species == "LTDU" ~ "Long-tailed Duck",
      Species == "STEI" ~ "Steller's Eider",
      Species == "SPEI" ~ "Spectacled Eider",
      TRUE ~ NA_character_   # drop anything unexpected
    ),
    Species = factor(
      Species,
      levels = c("King Eider", "Long-tailed Duck", "Steller's Eider", "Spectacled Eider")
    )
  ) %>%
  filter(!is.na(Species))

Fig2 <- ggplot(plot_dat, aes(x = day_since_init_int, y = n_nests, fill = Species)) +
  geom_col() +
  facet_wrap(~ Species, ncol = 2, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "King Eider"         = unname(palette["KIEI"]),
      "Long-tailed Duck"   = unname(palette["LTDU"]),
      "Steller's Eider"    = unname(palette["STEI"]),
      "Spectacled Eider"   = unname(palette["SPEI"])
    ),
    drop = FALSE
  ) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, NA)) +
  labs(
    x = "Days since nest initiation",
    y = "Number of nests monitored",
    fill = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )



Fig2
ggsave(
  "Fig2.png",
  plot = Fig2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 3.** Distribution of monitoring duration across the incubation period for four Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. Histograms show the number of nests monitored by day since nest initiation for *S. spectabilis*, *C. hyemalis*, *P. stelleri,* and *S. fischeri*, illustrating camera monitoring coverage and temporal distribution during incubation.

## Incubation Behavior

Incubation constancy defined as the **proportion of time spent on the nest, summarized as daily (all events per species summarized together) and total.**

**Table 3.** Mean values of incubation constancy with 95% confidence intervals for Arctic-breeding sea duck nests near Utqiagvik, Alaska, USA between 2011-2025.

```{r}
## data prep to exclude 'unviewable' nest-days 
hen_day <- hen.act.filtered %>%
  filter(Fate_simple %in% c("Hatched", "Failed")) %>%
  mutate(
    DateTimeBegin = ymd_hms(DateTimeBegin, tz = "UTC"),
    DateTimeEnd   = ymd_hms(DateTimeEnd,   tz = "UTC"),
    Date          = as.Date(DateTimeBegin),
    DurationMins  = as.numeric(difftime(DateTimeEnd, DateTimeBegin, units = "mins"))
  )

bad_nest_days <- hen_day %>%
  group_by(Species, Nest_ID, Date) %>%
  summarise(
    any_unviewable = any(Behavior == "U-Unviewable", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(any_unviewable) %>%
  select(Species, Nest_ID, Date)

hen_day_complete <- hen_day %>%
  anti_join(bad_nest_days, by = c("Species", "Nest_ID", "Date"))

## Incubation constancy (pooled by monitored minutes)
species_minutes <- hen_day_complete %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(Species, Fate_simple, Behavior) %>%
  summarise(total_minutes = sum(DurationMins, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Behavior, values_from = total_minutes, values_fill = 0) %>%
  mutate(
    total_monitored = `I-Incubating` + `N-NotIncubating`,
    constancy = if_else(total_monitored > 0, `I-Incubating` / total_monitored, NA_real_)
  )

# Overall per species, pooling all fates
species_all <- species_minutes %>%
  group_by(Species) %>%
  summarise(
    total_I = sum(`I-Incubating`, na.rm = TRUE),
    total_N = sum(`N-NotIncubating`, na.rm = TRUE),
    total_monitored = total_I + total_N,
    constancy = if_else(total_monitored > 0, total_I / total_monitored, NA_real_),
    .groups = "drop"
  )

## Recess duration per species (summarized across ALL recess events vs nest-avg)
recess_events <- hen_day_complete %>%
  filter(Behavior == "N-NotIncubating") %>%
  filter(!is.na(DurationMins), DurationMins >= 0)

recess_duration_summary_all <- recess_events %>%
  group_by(Species) %>%
  summarise(
    n_recess_events = n(),
    mean_recess_min = mean(DurationMins, na.rm = TRUE),
    sd_recess_min   = sd(DurationMins, na.rm = TRUE),
    median_recess_min = median(DurationMins, na.rm = TRUE),
    IQR_recess_min    = IQR(DurationMins, na.rm = TRUE),
    .groups = "drop"
  )

# duration by summary by fate option
recess_duration_summary_by_fate <- recess_events %>%
  group_by(Species, Fate_simple) %>%
  summarise(
    n_recess_events = n(),
    mean_recess_min = mean(DurationMins, na.rm = TRUE),
    sd_recess_min   = sd(DurationMins, na.rm = TRUE),
    median_recess_min = median(DurationMins, na.rm = TRUE),
    IQR_recess_min    = IQR(DurationMins, na.rm = TRUE),
    .groups = "drop"
  )

## Recesses per monitored day INCLUDING zeros
# build set of monitored nest-days after removing Unviewable days

monitored_nest_days <- hen_day_complete %>%
  distinct(Species, Fate_simple, Nest_ID, Date)

# Count recess events per monitored nest-day (days with none will be missing -> fill to 0)
recesses_per_day <- recess_events %>%
  group_by(Species, Fate_simple, Nest_ID, Date) %>%
  summarise(recess_count = n(), .groups = "drop")

recess_day_full <- monitored_nest_days %>%
  left_join(recesses_per_day, by = c("Species", "Fate_simple", "Nest_ID", "Date")) %>%
  mutate(recess_count = tidyr::replace_na(recess_count, 0L))

# summary across all (day-weighted, including zeros)
recess_freq_summary_all <- recess_day_full %>%
  group_by(Species) %>%
  summarise(
    n_monitored_days = n(),
    mean_recesses_per_day = mean(recess_count, na.rm = TRUE),
    sd_recesses_per_day   = sd(recess_count, na.rm = TRUE),
    median_recesses_per_day = median(recess_count, na.rm = TRUE),
    IQR_recesses_per_day    = IQR(recess_count, na.rm = TRUE),
    .groups = "drop"
  )

# recess frequency by fate
recess_freq_summary_by_fate <- recess_day_full %>%
  group_by(Species, Fate_simple) %>%
  summarise(
    n_monitored_days = n(),
    mean_recesses_per_day = mean(recess_count, na.rm = TRUE),
    sd_recesses_per_day   = sd(recess_count, na.rm = TRUE),
    median_recesses_per_day = median(recess_count, na.rm = TRUE),
    IQR_recesses_per_day    = IQR(recess_count, na.rm = TRUE),
    .groups = "drop"
  )

# combine into one summary
species_behavior_summary <- species_all %>%
  left_join(recess_freq_summary_all, by = "Species") %>%
  left_join(recess_duration_summary_all, by = "Species")

```

Constancy was computed from pooled minutes per species-day; summaries reflect variation among days. Sample sizes report unique nests.

```{r}
# Pub quality figure

# Nest-level incubation constancy, proportional
nest_const <- hen_day_complete %>%
  filter(Fate_simple %in% c("Hatched", "Failed")) %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(Species, Nest_ID, Fate_simple, Behavior) %>%
  summarise(total_minutes = sum(DurationMins, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Behavior, values_from = total_minutes, values_fill = 0) %>%
  mutate(
    total_monitored = `I-Incubating` + `N-NotIncubating`,
    constancy = if_else(total_monitored > 0, `I-Incubating` / total_monitored, NA_real_)
  ) %>%
  filter(!is.na(constancy))

# Species by fate summary with 95% CI
summ_sf <- nest_const %>%
  group_by(Species, Fate_simple) %>%
  summarise(
    n_nests = n_distinct(Nest_ID),
    mean_const = mean(constancy, na.rm = TRUE),
    sd_const   = sd(constancy, na.rm = TRUE),
    se_const   = sd_const / sqrt(n_nests),
    tcrit      = qt(0.975, df = pmax(n_nests - 1, 1)),
    lo         = mean_const - tcrit * se_const,
    hi         = mean_const + tcrit * se_const,
    .groups = "drop"
  ) %>%
  mutate(
    # clamp to biological bounds
    lo = pmax(lo, 0),
    hi = pmin(hi, 1),

    # convert to %
    mean_pct = 100 * mean_const,
    lo_pct   = 100 * lo,
    hi_pct   = 100 * hi,

    # relabel species to common names
    Species_common = dplyr::recode(as.character(Species), !!!species_labels),
    Species_common = factor(Species_common, levels = unname(species_labels))
  )

# Species/Fate/N/Mean/CI

summ_sf %>%
  transmute(
    Species = Species_common,
    Fate = Fate_simple,
    `N (nests)` = n_nests,
    `Mean Incubation Constancy (%)` = round(mean_pct, 1),
    `Lower 95% CI` = round(lo_pct, 1),
    `Upper 95% CI` = round(hi_pct, 1)
  ) %>%
  gt() %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )
```

**Recess Behavior Summary Statistics**

```{r}
## Constancy by species and fate, minutes pooled
constancy_by_fate <- species_minutes %>%
  group_by(Species, Fate_simple) %>%
  summarise(
    total_I = sum(`I-Incubating`, na.rm = TRUE),
    total_N = sum(`N-NotIncubating`, na.rm = TRUE),
    total_monitored = total_I + total_N,
    constancy = if_else(total_monitored > 0, total_I / total_monitored, NA_real_),
    .groups = "drop"
  )

# Nest counts by species and fate
nests_by_fate <- hen_day_complete %>%
  distinct(Species, Fate_simple, Nest_ID) %>%
  count(Species, Fate_simple, name = "n_nests")

# combine to one fate table
species_behavior_by_fate <- constancy_by_fate %>%
  left_join(nests_by_fate, by = c("Species", "Fate_simple")) %>%
  left_join(recess_freq_summary_by_fate, by = c("Species", "Fate_simple")) %>%
  left_join(recess_duration_summary_by_fate, by = c("Species", "Fate_simple")) %>%
  mutate(
    constancy_pct = 100 * constancy
  ) %>%
  arrange(Species, Fate_simple)

# build table
species_behavior_by_fate %>%
  dplyr::select(
    Species,
    Fate_simple,
    n_nests,
    n_monitored_days,
    `Incubation constancy (%)` = constancy_pct,
    `Mean recesses/day` = mean_recesses_per_day,
    `SD recesses/day` = sd_recesses_per_day,
    n_recess_events,
    `Mean recess duration (min)` = mean_recess_min,
    `SD recess duration (min)` = sd_recess_min,
    `Median recess duration (min)` = median_recess_min,
    `IQR recess duration (min)` = IQR_recess_min
  ) %>%
  knitr::kable(format = "html", digits = 2, align = "l",
              caption = "Incubation constancy, recess frequency, and recess duration summarized by species and nest fate (Hatched vs Failed).") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 12
  )

summ_sf %>%
  transmute(
    Species = Species_common,
    Fate = Fate_simple,
    `N (nests)` = n_nests,
    `Mean Incubation Constancy (%)` = round(mean_pct, 1),
    `Lower 95% CI` = round(lo_pct, 1),
    `Upper 95% CI` = round(hi_pct, 1)
  ) %>%
  gt() %>%
  cols_align(
    align = "left",
    columns = everything()
  ) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )

species_behavior_summary %>%
  mutate(
    Species = dplyr::recode(as.character(Species), !!!species_labels)
  ) %>%
  dplyr::select(
    Species,
    n_monitored_days,
    n_recess_events,
    `Incubation constancy (%)` = constancy,
    `Mean recesses/day`        = mean_recesses_per_day,
    `SD recesses/day`          = sd_recesses_per_day,
    `Mean recess duration (min)`   = mean_recess_min,
    `SD recess duration (min)`     = sd_recess_min,
    `Median recess duration (min)` = median_recess_min,
    `IQR recess duration (min)`    = IQR_recess_min
  ) %>%
  mutate(
    `Incubation constancy (%)` = round(100 * `Incubation constancy (%)`, 1),
    across(where(is.numeric), ~ round(.x, 2))
  ) %>%
  gt() %>%
  cols_align(align = "left", columns = everything()) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )


# species level grouped incubation constancy 
nest_const_all <- hen_day_complete %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(Species, Nest_ID, Behavior) %>%
  summarise(total_minutes = sum(DurationMins, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from  = Behavior,
    values_from = total_minutes,
    values_fill = 0
  ) %>%
  mutate(
    total_monitored = `I-Incubating` + `N-NotIncubating`,
    constancy = dplyr::if_else(total_monitored > 0, `I-Incubating` / total_monitored, NA_real_)
  ) %>%
  filter(!is.na(constancy))

summ_species <- nest_const_all %>%
  group_by(Species) %>%
  summarise(
    n_nests = n_distinct(Nest_ID),
    mean_const = mean(constancy, na.rm = TRUE),
    sd_const   = sd(constancy, na.rm = TRUE),
    se_const   = sd_const / sqrt(n_nests),
    tcrit      = qt(0.975, df = pmax(n_nests - 1, 1)),
    lo         = pmax(mean_const - tcrit * se_const, 0),
    hi         = pmin(mean_const + tcrit * se_const, 1),
    .groups = "drop"
  ) %>%
  mutate(
    Species_common = dplyr::recode(as.character(Species), !!!species_labels),
    mean_pct = 100 * mean_const,
    lo_pct   = 100 * lo,
    hi_pct   = 100 * hi
  )

gt_tbl <- summ_species %>%
  transmute(
    Species = Species_common,
    `N (nests)` = n_nests,
    `Mean Incubation Constancy (%)` = round(mean_pct, 1),
    `Lower 95% CI` = round(lo_pct, 1),
    `Upper 95% CI` = round(hi_pct, 1)
  ) %>%
  gt() %>%
  cols_align(align = "left", columns = everything()) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )

gt_tbl
```

```{r}
# Nest-level incubation constancy (reminder, hen_day_complete already excludes incomplete nest days)

nest_behavior_summary <- hen_day_complete %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(Species, Nest_ID, Fate_simple, Behavior) %>%
  summarise(
    total_minutes = sum(DurationMins, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from  = Behavior,
    values_from = total_minutes,
    values_fill = 0
  ) %>%
  mutate(
    total_monitored = `I-Incubating` + `N-NotIncubating`,
    constancy = dplyr::if_else(total_monitored > 0, `I-Incubating` / total_monitored, NA_real_)
  ) %>%
  filter(!is.na(constancy))

# Extract nest-level constancy plus fate (this now matches upstream filtering)
nest_df <- nest_behavior_summary %>%
  filter(Fate_simple %in% c("Hatched", "Failed")) %>%
  select(Species, Nest_ID, Fate_simple, constancy)

#  prep plot dataframe
df <- nest_df %>%
  mutate(
    Species = dplyr::recode(as.character(Species), !!!species_labels),
    Species = factor(Species, levels = unname(species_labels)),
    Fate_simple = factor(Fate_simple, levels = c("Failed", "Hatched"))
  ) %>%
  filter(!is.na(constancy), !is.na(Species), !is.na(Fate_simple))

# group summaries (mean ± 95% CI)
sum_sf <- df %>%
  group_by(Species, Fate_simple) %>%
  summarise(
    n    = n(),
    mean = mean(constancy, na.rm = TRUE),
    sd   = sd(constancy, na.rm = TRUE),
    se   = sd / sqrt(n),
    tcrit = qt(0.975, df = pmax(n - 1, 1)),
    lo   = mean - tcrit * se,
    hi   = mean + tcrit * se,
    .groups = "drop"
  )

# Wilcoxon within species and labels
p_lab <- df %>%
  group_by(Species) %>%
  summarise(
    p_value = if_else(
      all(c("Failed", "Hatched") %in% Fate_simple) &
        sum(Fate_simple == "Failed")  > 0 &
        sum(Fate_simple == "Hatched") > 0,
      wilcox.test(constancy ~ Fate_simple, exact = FALSE)$p.value,
      NA_real_
    ),
    y_pos = max(constancy, na.rm = TRUE) + 0.05 * diff(range(constancy, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    label = case_when(
      is.na(p_value)   ~ "p = NA",
      p_value < 0.001  ~ "p < 0.001",
      TRUE             ~ paste0("p = ", signif(p_value, 2))
    )
  )

p_lab <- p_lab %>%
  mutate(y_pos = pmin(y_pos, 0.98))

# plot updated figure (w/o box and violin plots)
Fig3 <- ggplot(df, aes(x = Fate_simple, y = constancy, color = Species)) +
  geom_errorbar(
    data = sum_sf,
    aes(x = Fate_simple, ymin = lo, ymax = hi, color = Species),
    width = 0.18,
    linewidth = 1.1,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = p_lab,
    aes(x = 1.5, y = y_pos, label = label),
    inherit.aes = FALSE,
    size = 4,
    color = "grey20"
  ) +
  facet_wrap(~ Species, ncol = 2) +   # <- removed scales = "free_y"
  scale_color_manual(values = pal_common, guide = "none") +
  coord_cartesian(ylim = c(0.25, 1.00)) +   # <- fixed y limits
  labs(
    x = "Nest fate",
    y = "Incubation constancy"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )

Fig3

# export as PNG
ggsave(
  "Fig3.png",
  plot = Fig3,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 3.** Mean daily incubation constancy by nest fate for four Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. Points represent nest-level mean incubation constancy (excluding the terminal day), with larger points and error bars indicating species-specific means ± SD for failed and hatched nests. P-values denote results of species-specific comparisons between nest fate categories.

```{r}
#| warning: false
# establish stages
stage_breaks <- c(-0.5, 3.5, 7.5, 11.5, 15.5, 19.5, 23.5, 27.5)
stage_labels <- c("0–3", "4–7", "8–11", "12–15", "16–19", "20–23", "24–27")
dodge_w <- 0.65

add_stage <- function(df) {
  df%>%
    mutate(
      day_since_init_int = floor(day_since_init),
      stage = cut(
        day_since_init_int,
        breaks = stage_breaks,
        labels = stage_labels,
        right  = TRUE
      ),
      stage = factor(stage, levels = stage_labels, ordered = TRUE)
    ) %>%
    filter(day_since_init_int >= 0, day_since_init_int <= 27, !is.na(stage))
}

# summarizer: mean ± SE (NOT CI)
summarise_se <- function(df, value_col) {
  value_col <- rlang::ensym(value_col)
  df %>%
    summarise(
      n    = sum(!is.na(!!value_col)),
      mean = mean(!!value_col, na.rm = TRUE),
      sd   = sd(!!value_col, na.rm = TRUE),
      se   = sd / sqrt(n),
      lo   = mean - se,
      hi   = mean + se,
      .groups = "drop"
    )
}

# IC (nest-level, stage-level mean ± SE)
hen_daily <- hen.act.filtered %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(Species, Nest_ID, day_since_init_int = floor(day_since_init)) %>%
  summarise(
    mins_inc = sum(DurationMins[Behavior == "I-Incubating"], na.rm = TRUE),
    mins_non = sum(DurationMins[Behavior == "N-NotIncubating"], na.rm = TRUE),
    mins_tot = mins_inc + mins_non,
    constancy = if_else(mins_tot > 0, mins_inc / mins_tot, NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    stage = cut(day_since_init_int, breaks = stage_breaks, labels = stage_labels, right = TRUE),
    stage = factor(stage, levels = stage_labels, ordered = TRUE)
  ) %>%
  filter(day_since_init_int >= 0, day_since_init_int <= 27, !is.na(stage))

hen_stage_nest <- hen_daily %>%
  group_by(Species, Nest_ID, stage) %>%
  summarise(constancy = mean(constancy, na.rm = TRUE), .groups = "drop")

stage_summ <- hen_stage_nest %>%
  group_by(Species, stage) %>%
  group_modify(~ summarise_se(.x, constancy)) %>%
  ungroup() %>%
  transmute(
    Species, stage,
    n_nests    = n,
    mean_const = mean * 100,
    se_low     = lo * 100,
    se_high    = hi * 100
  )

# table of sample sizes
table_stage <- stage_summ %>%
  select(Species, stage, n_nests) %>%
  pivot_wider(names_from = stage, values_from = n_nests) %>%
  arrange(Species)

table_grob <- tableGrob(
  table_stage,
  rows = NULL,
  theme = ttheme_minimal(
    base_size = 9,
    core = list(bg_params = list(fill = NA))
  )
)

# recess events
recess_events <- hen.act.filtered %>%
  filter(Behavior == "N-NotIncubating") %>%
  add_stage()

# Recess frequency (mean daily recesses per nest per stage, ±SE)
recess_count_sum <- recess_events %>%
  group_by(Species, Nest_ID, stage) %>%
  summarise(recesses = n(), .groups = "drop") %>%
  group_by(Species, stage) %>%
  group_modify(~ summarise_se(.x, recesses)) %>%
  ungroup() %>%
  transmute(
    Species, stage,
    mean_cnt = mean,
    se_low   = lo,
    se_high  = hi
  )

# recess duration (mean recess duration per nest per stage, ±SE)
recess_dur_sum <- recess_events %>%
  group_by(Species, Nest_ID, stage) %>%
  summarise(mean_recess_min = mean(DurationMins, na.rm = TRUE), .groups = "drop") %>%
  group_by(Species, stage) %>%
  group_modify(~ summarise_se(.x, mean_recess_min)) %>%
  ungroup() %>%
  transmute(
    Species, stage,
    mean_dur = mean,
    se_low   = lo,
    se_high  = hi
  )

# style
base_theme <- theme_bw(base_size = 15) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )

# figs
fig_incubation <- ggplot(
  stage_summ,
  aes(x = stage, y = mean_const, color = Species, shape = Species)
) +
  geom_errorbar(aes(ymin = se_low, ymax = se_high),
                width = 0.2, linewidth = 0.8,
                position = position_dodge(width = dodge_w)) +
  geom_point(size = 3.5, stroke = 1,
             position = position_dodge(width = dodge_w)) +
  scale_color_manual(values = palette) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_x_discrete(name = "Incubation stage (days)") +
  scale_y_continuous(name = "Constancy (%)",
                     expand = expansion(mult = c(0.02, 0.02))) +
  coord_cartesian(ylim = c(0, 100)) +
  base_theme

fig_recess_count <- ggplot(
  recess_count_sum,
  aes(x = stage, y = mean_cnt, color = Species, shape = Species)
) +
  geom_errorbar(aes(ymin = se_low, ymax = se_high),
                width = 0.2, linewidth = 0.8,
                position = position_dodge(width = dodge_w)) +
  geom_point(size = 3.5, stroke = 1,
             position = position_dodge(width = dodge_w)) +
  scale_color_manual(values = palette) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_x_discrete(name = "Incubation stage (days)") +
  scale_y_continuous(name = "Number of daily recesses",
                     expand = expansion(mult = c(0.02, 0.02))) +
  coord_cartesian(ylim = c(0, NA)) +
  base_theme

fig_recess_duration <- ggplot(
  recess_dur_sum,
  aes(x = stage, y = mean_dur, color = Species, shape = Species)
) +
  geom_errorbar(aes(ymin = se_low, ymax = se_high),
                width = 0.2, linewidth = 0.8,
                position = position_dodge(width = dodge_w)) +
  geom_point(size = 3.5, stroke = 1,
             position = position_dodge(width = dodge_w)) +
  scale_color_manual(values = palette) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_x_discrete(name = "Incubation stage (days)") +
  scale_y_continuous(name = "Mean recess duration (min)",
                     expand = expansion(mult = c(0.02, 0.02))) +
  coord_cartesian(ylim = c(0, 250)) +
  base_theme

# print outputs
fig_incubation
fig_recess_count
fig_recess_duration
table_stage

```

**Figure EXCLUDE.** Incubation constancy, recess frequency, and recess duration across 4-day incubation stages for four sea duck species. Points show nest-level means averaged within stages, and error bars represent 95% confidence intervals, calculated from time-lapse video monitoring. Large and not biologically realistic confidence intervals are due to low sample sizes.

**Hourly recess behavior**

```{r}
core_frac <- 0.50  # dominant window contains this fraction of recesses (try 0.5–0.7)

# one row per recess w/ hour (COMMON NAMES)
recess_tod <- hen.act.filtered %>%
  filter(Behavior == "N-NotIncubating") %>%
  mutate(
    DateTimeBegin = ymd_hms(DateTimeBegin, tz = "UTC"),
    hour = lubridate::hour(DateTimeBegin),
    Species = dplyr::recode(as.character(Species), !!!species_labels),
    Species = factor(Species, levels = unname(species_labels))
  ) %>%
  filter(!is.na(hour), !is.na(Species)) %>%
  distinct(Nest_ID, Species, ID, hour)  # one per recess event

# Counts by hour
hour_counts <- recess_tod %>%
  count(Species, hour, name = "n_recesses") %>%
  # ensure all hours exist for each species (so window-finding is stable)
  tidyr::complete(Species, hour = 0:23, fill = list(n_recesses = 0L))

Fig4_count <- ggplot(hour_counts, aes(x = hour, y = n_recesses, fill = Species)) +
  geom_col(show.legend = FALSE, width = 0.85) +
  facet_wrap(~ Species, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = pal_common, drop = FALSE) +
  scale_x_continuous(breaks = seq(0, 23, 3), minor_breaks = NULL) +
  labs(
    x = "Hour of day (recess start time)",
    y = "Number of recesses",
    title = "Starting hour of recesses (counts)"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

Fig4_count

# proportion by hour within species
hour_summ <- hour_counts %>%
  group_by(Species) %>%
  mutate(prop_recesses = n_recesses / sum(n_recesses, na.rm = TRUE)) %>%
  ungroup()

# --- helper: find smallest contiguous (circular) window containing core_frac of recesses
find_core_window <- function(hour, prop, frac = 0.5) {
  full <- tibble(hour = 0:23) %>%
    left_join(tibble(hour = hour, prop = prop), by = "hour") %>%
    mutate(prop = replace_na(prop, 0)) %>%
    arrange(hour)

  # duplicate for circular wrap
  h2 <- c(full$hour, full$hour + 24)
  p2 <- c(full$prop, full$prop)

  best_len <- Inf
  best_i <- best_j <- NA
  i <- 1
  s <- 0

  for (j in seq_along(p2)) {
    s <- s + p2[j]
    while (s - p2[i] >= frac) {
      s <- s - p2[i]
      i <- i + 1
    }
    # start must be within first day; window length cannot exceed 24
    if (s >= frac && h2[i] < 24 && h2[j] <= h2[i] + 24) {
      len <- h2[j] - h2[i]
      if (len < best_len) {
        best_len <- len
        best_i <- i
        best_j <- j
      }
    }
  }

  xmin <- h2[best_i]
  xmax <- h2[best_j] + 1  # +1 so last hour bin is visually included

  # if crosses midnight, return two rectangles
  if (xmax <= 24) {
    tibble(xmin = xmin, xmax = xmax)
  } else {
    bind_rows(
      tibble(xmin = xmin, xmax = 24),
      tibble(xmin = 0,    xmax = xmax - 24)
    )
  }
}

# build per-species shading rectangles
shade_df <- hour_summ %>%
  group_by(Species) %>%
  summarise(rects = list(find_core_window(hour, prop_recesses, frac = core_frac)),
            .groups = "drop") %>%
  unnest(rects) %>%
  mutate(ymin = -Inf, ymax = Inf)

# sample size recesses
n_recess_labels <- hour_counts %>%
  group_by(Species) %>%
  summarise(
    n_recess = sum(n_recesses),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("n = ", n_recess)
  )

# Plot proportions + dominant window shading (species-specific)
Fig4 <- ggplot(hour_summ, aes(x = hour, y = prop_recesses, fill = Species)) +
  # dominant window shading
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey30",
    alpha = 0.12
  ) +
  geom_col(show.legend = FALSE, width = 0.85) +
  facet_wrap(~ Species, ncol = 2) +
  scale_fill_manual(values = pal_common, drop = FALSE) +
  scale_x_continuous(breaks = seq(0, 23, 3), minor_breaks = NULL) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = NULL,
    x = "Hour of day (recess start time)",
    y = "Proportion of recesses"
  ) +
  # sample size annotation
  geom_text(
    data = n_recess_labels,
    aes(x = 23.5, y = 0.125, label = label),
    inherit.aes = FALSE,
    hjust = 1,
    vjust = 1,
    size = 4
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.4),
    strip.text = element_text(face = "bold")
  )

Fig4

ggsave(
  "Fig4.png",
  plot = Fig4,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 4.** Distribution of incubation recess start times across the diel cycle for four Arctic-breeding sea ducks nears Utqiaġvik, Alaska between 2011-2025. Bars show the proportion of observed recesses initiating in each hour of the day. Sample sizes (n) indicate the total number of recess per species. Shaded regions denote the smallest contiguous time interval containing 50% of recess starts.

## Daily Incubation Constancy Analysis

We evaluated seasonal, environmental, and species-specific drivers of daily incubation constancy using linear mixed-effects models. Analyses were conducted at the nest-day level, with one observation per nest per day of monitoring. Because incubation constancy is a bounded proportion (0-1) and frequently approached the upper boundary, we modeled daily constancy on the logit scale, and included a boundary adjustment (ε = 0.001) to prevent infinite values. Nest identity was included as a random intercept to account for repeated daily observations within nests. Year was initially included as a random intercept to account for interannual variation, but was removed from the final models when variance estimates were consistently near zero (add estimates), which indicated it contributed negligible explanatory power after nest-level effects were accounted for.

Fixed effects included species identity, day of the year (DOY) of recess event to represent seasonal timing, days since incubation initiation to capture within-nest energetic progression, and daily weather variables. We included daily maximum air temperature (TMAX) and mean daily wind speed (WND) as weather predictors based on prior analyses indicating their influence on waterfowl incubation behaviors (refs). Daily weather data was retrieved from the NOAA meteorological station at Point Barrow (waypoints; ref) and z-transformed to ease interpretation and reduce collinearity. We evaluated a candidate set of models that varied in their inclusion of seasonal, weather, and species-specific effects, including one intercept only model, and models with Species x DOY interactions to test whether seasonal trajectories differed among species. Models were compared using Akaike’s Information Criterion corrected for small sample size (AICc), and inference focused on relative support among competing hypotheses rather than reliance on a single full model. This modeling framework allowed us to distinguish between broad seasonal effects associated with Arctic phenology and finer-scale within-nest progression through incubation.

```{r}
# Logit-normal mixed models for daily incubation constancy
# Rresponse: proportional daily incubation constancy (0–1)
# Model:    lmer() on logit-transformed proportion (logit-normal)
# Random:   Nest_ID and Year, if supported by variance estimates. 

# data prep
daily_ducks <- hen.act.filtered %>%
  filter(Behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  mutate(
    DateTimeBegin = ymd_hms(DateTimeBegin),
    date = as.Date(DateTimeBegin)
  ) %>%
  group_by(Species, Nest_ID, date) %>%
  summarise(
    mins_inc  = sum(DurationMins[Behavior == "I-Incubating"], na.rm = TRUE),
    mins_non  = sum(DurationMins[Behavior == "N-NotIncubating"], na.rm = TRUE),
    mins_tot  = mins_inc + mins_non,
    constancy_daily = if_else(mins_tot > 0, mins_inc / mins_tot, NA_real_),
    recesses_daily  = sum(Behavior == "N-NotIncubating", na.rm = TRUE),
    day_since_init  = mean(day_since_init, na.rm = TRUE),
    .groups = "drop"
  )

## add in daily weather variables (TMAX + wind)
# drop TMIN due to correlation with TMAX
met_daily <- met_raw %>%
  mutate(
    date    = as.Date(DATE),
    TMAX_C  = ifelse(is.na(TMAX), NA_real_, TMAX / 10),
    AWND_ms = ifelse(is.na(AWND), NA_real_, AWND / 10)
  ) %>%
  select(date, TMAX_C, AWND_ms) %>%
  arrange(date)

daily_wx <- daily_ducks %>%
  left_join(met_daily, by = "date") %>%
  mutate(
    date    = ymd(date),
    DOY     = yday(date),
    Year    = factor(year(date)),
    Species = factor(Species),
    Nest_ID = factor(Nest_ID)
  )

# standardize analysis dataset, add boundary adjustment eps for 0/1 values
constancy_camera <- daily_wx %>%
  filter(
    !is.na(constancy_daily),
    !is.na(DOY),
    !is.na(Year),
    !is.na(TMAX_C),
    !is.na(AWND_ms),
    is.finite(day_since_init),
    mins_tot > 0
  ) %>%
  mutate(
    # z-scales for interpretability/convergence
    z_DOY   = as.numeric(scale(DOY)),
    z_TMAX  = as.numeric(scale(TMAX_C)),
    z_AWND  = as.numeric(scale(AWND_ms)),
    z_stage = as.numeric(scale(day_since_init)),
    eps = 0.001, # boundary adjustment
    constancy_adj = pmin(pmax(constancy_daily, eps), 1 - eps),
    logit_const   = qlogis(constancy_adj)
  )

# Model fitting! 
ctrl <- lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5)
)

# add a random effect rule for Nest_ID and Year for updated models. 
# If Year routinely goes singular in your data, keep only (1 | Nest_ID).
re_form <- "(1 | Nest_ID) + (1 | Year)"  # change to "(1 | Nest_ID)" if Year is singular

m_cam_null <- lmer(
  as.formula(paste("logit_const ~ 1 +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_stage <- lmer(
  as.formula(paste("logit_const ~ z_stage + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_doy <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_TMAX <- lmer(
  as.formula(paste("logit_const ~ z_stage + z_TMAX + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_AWND <- lmer(
  as.formula(paste("logit_const ~ z_stage + z_AWND + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_TMAX_AWND <- lmer(
  as.formula(paste("logit_const ~ z_stage + z_TMAX + z_AWND + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_doy_TMAX <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + z_TMAX + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_doy_AWND <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + z_AWND + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_doy_TMAX_AWND <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + z_TMAX + z_AWND + Species +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

m_cam_int_doy <- lmer(
  as.formula(paste("logit_const ~ z_DOY * Species + z_stage + z_TMAX + z_AWND +", re_form)),
  data    = constancy_camera,
  control = ctrl,
  REML    = FALSE
)

## Make AIC TABLE
cam_set <- list(
  null              = m_cam_null,
  stage             = m_cam_stage,
  doy               = m_cam_doy,
  TMAX              = m_cam_TMAX,
  wind              = m_cam_AWND,
  TMAX_wind         = m_cam_TMAX_AWND,
  doy_TMAX          = m_cam_doy_TMAX,
  doy_wind          = m_cam_doy_AWND,
  doy_TMAX_wind     = m_cam_doy_TMAX_AWND,
  int_doy_TMAX_wind = m_cam_int_doy
)

model_labels <- c(
  null              = "Intercept (random Nest + Year)",
  stage             = "Stage",
  doy               = "DOY + stage",
  TMAX              = "Stage + TMAX",
  wind              = "Stage + wind",
  TMAX_wind         = "Stage + TMAX + wind",
  doy_TMAX          = "DOY + stage + TMAX",
  doy_wind          = "DOY + stage + wind",
  doy_TMAX_wind     = "DOY + stage + TMAX + wind",
  int_doy_TMAX_wind = "DOY × Species + stage + TMAX + wind"
)

# helper function for rows
get_nobs <- function(m) {
  # works for lmer; falls back gracefully
  nobs(m)
}

aicc_cam_tbl <- purrr::imap_dfr(cam_set, \(m, nm) {
  ll <- logLik(m)
  tibble(
    Model_key = nm,
    n         = get_nobs(m),
    K         = attr(ll, "df"),
    logLik    = as.numeric(ll),
    AIC       = AIC(m),
    AICc      = MuMIn::AICc(m)
  )
}) %>%
  mutate(Model = dplyr::recode(Model_key, !!!model_labels)) %>%
  arrange(AICc) %>%
  mutate(
    `ΔAICc` = AICc - min(AICc, na.rm = TRUE),
    Weight  = exp(-0.5 * `ΔAICc`) / sum(exp(-0.5 * `ΔAICc`), na.rm = TRUE),
    Weight  = round(Weight, 3)
  )

# pull best model via AICc - assign best fit model manually depending on table context. 
best_key <- aicc_cam_tbl$Model_key[1]
m_best   <- cam_set[[best_key]]


# Publication quality table
Table4 <- aicc_cam_tbl %>%
  select(
    Model,
    K,
    logLik,
    AICc,
    `ΔAICc`,
    Weight
  ) %>%
  gt() %>%
  cols_align(align = "center", columns = c(K, logLik, AICc, `ΔAICc`, Weight)) %>%
  cols_align(align = "left", columns = Model) %>%
  fmt_number(columns = "K", decimals = 0) %>%
  fmt_number(columns = c("logLik", "AICc", `ΔAICc`), decimals = 2) %>%
  fmt_number(columns = "Weight", decimals = 3) %>%
  cols_label(
    Model   = "Model",
    K       = "K",
    logLik  = html("logLik"),
    AICc    = html("AIC<sub>c</sub>"),
    `ΔAICc` = html("&Delta;AIC<sub>c</sub>"),
    Weight  = html("w<sub>i</sub>")
  ) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )

Table4

# export table for manuscript
gtsave(
  Table4,
  filename = "Table4.png",
  vwidth  = 900,
  vheight = 300
)
```

**Table 4.** Model selection results for factors affecting daily incubation constancy of Arctic-breeding sea duck nests in Utqiagivk, Alaska between 2011-2025. All models include a random intercept for nest identity and year.

**Best fit model summary and diagnostics**

```{r}
summary(m_best)
AIC(m_best)
logLik(m_best)

### Residual diagnostics
resid_raw   <- residuals(m_best)
resid_pear  <- residuals(m_best, type="pearson")
fitted_vals <- fitted(m_best)

# Residuals vs fitted
plot(fitted_vals, resid_raw,
     xlab= "Fitted values (logit)", ylab ="Residuals",
     main="Residuals vs fitted"); abline(h=0, lty=2)

# Normal Q-Q plot
qqnorm(resid_raw); qqline(resid_raw, lty=2)

# Histogram of residuals
hist(resid_raw, breaks=40, main="Histogram of residuals", xlab="Residual (logit scale)")

# Scale-location (variance vs mean) – optional quick check
plot(fitted_vals, sqrt(abs(resid_raw)),
     xlab= "Fitted values (logit scale)", ylab="sqrt(|residual|)",
     main="Scale-location"); lines(lowess(fitted_vals, sqrt(abs(resid_raw))), col=2)
#Primary figure: Predicted daily IC vs DOY, by species
inv_logit <- function(x) plogis(x)


# build prediction grid

```

```{r}
doy_seq <- seq(
  min(constancy_camera$DOY, na.rm = TRUE),
  max(constancy_camera$DOY, na.rm = TRUE),
  length.out = 200
)
pred_grid <- expand.grid(
  DOY     = doy_seq,
  Species = levels(constancy_camera$Species)
) %>%
  as_tibble() %>%
  mutate(
    # recreate z_DOY using same mean/sd as analysis data
    z_DOY = (DOY - mean(constancy_camera$DOY, na.rm = TRUE)) /
      sd(constancy_camera$DOY, na.rm = TRUE),

    # hold other standardized covariates at their means (0 if scaled)
    z_stage = 0,
    z_TMAX  = 0,
    z_AWND  = 0
  )

# Fixed-effect predictions, and SEs on link scale
# Use the fixed-effects design matrix and variance-covariance matrix
X <- model.matrix(
  delete.response(terms(m_best)),
  pred_grid
)

b <- fixef(m_best)
V <- vcov(m_best)

# Align columns
X <- X[, names(b), drop = FALSE]

fit_link <- as.vector(X %*% b)
se_link  <- sqrt(pmax(0, diag(X %*% V %*% t(X))))

pred_grid <- pred_grid %>%
  mutate(
    fit_link = fit_link,
    se_link  = se_link,
    lo_link  = fit_link - 1.96 * se_link,
    hi_link  = fit_link + 1.96 * se_link,
    fit = inv_logit(fit_link),
    lo  = inv_logit(lo_link),
    hi  = inv_logit(hi_link)
  )

# Relabel species with common names and assign palette. 
pred_plot <- pred_grid %>%
  mutate(
    Species = recode_factor(Species, !!!species_labels)
  )

# prep data for x-axis rugs
raw_plot <- constancy_camera %>%
  dplyr::filter(!is.na(DOY), !is.na(Species)) %>%
  mutate(Species = recode_factor(Species, !!!species_labels))

# Pub-quality figure
Fig_primary <- ggplot() +
  geom_ribbon(
    data = pred_plot,
    aes(x = DOY, ymin = lo, ymax = hi, fill = Species),
    alpha = 0.25,
    linewidth = 0
  ) +
  geom_rug(
  data = raw_plot,
  aes(x = DOY, color = Species),
  sides = "b",
  alpha = 0.2,
  inherit.aes = FALSE
) +
  # prediction lines
  geom_line(
    data = pred_plot,
    aes(x = DOY, y = fit, color = Species),
    linewidth = 1.4
  ) +
  facet_wrap(~ Species, ncol = 2) +
  scale_color_manual(values = pal_common, guide = "none") +
  scale_fill_manual(values = pal_common, guide = "none") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0.60, 1.00),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  labs(
    x = "Day of year (DOY)",
    y = "Predicted daily incubation constancy"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.4, color = "grey80"),
    panel.spacing      = unit(0.9, "lines"),
    strip.text         = element_text(face = "bold")
  )

Fig_primary

ggsave(
  "Fig5.png",
  plot = Fig_primary,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 5.** Predicted daily incubation constancy as a function of day of the year for Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. Curves represent population-level predictions from the top-ranked general mixed models including a day of year and species interaction, holding incubation stage and weather covariates at their mean values. Shaded bands indicate 95% confidence intervals. Along the x-axis, days range from 170 (June 19) to 210 (July 29), and rugs show the distribution of observed nest-days for each species.

**Key Model Takeaways**

-   Strong support for seasonal effects model that varies by species (DOY x Species + stage + TMAX + wind).

-   Seasonality (day of year) and nest stage are the main drivers.

-   Early vs. late incubation is important, no matter how early or late nest is initiated. Likely reflects shifts in female body condition, predator encounter risk..

-   Short-term weather (that we are able to measure) plays a lesser role in daily incubation constancy behavior.

-   KIEI and SPEIs maintained high constancy early in the season, followed by marked declines late in incubation, whereas LTDU showed a more gradual seasonal decrease and STEI relatively stable constancy across the season. Consistent with large-bodied vs small-bodied duck hypothesis.

## Daily Nest Survival - Logistic Exposure Models

We evaluated the relationships between incubation behavior, seasonal timing and nest fate using logistic-exposure models which account for variable monitoring intervals and allow estimation of daily nest survival probabilities (Schaffer 2004). Each nest contributed a series of exposure intervals defined by consecutive observation days, terminating at hatch or failure. Because each nest experience at most one failure event, we did not include nest-level random effects to avoid overparameterization. The response variable was survival to the next observation interval (1 = survived, 0 = failed on the final interval), modeled using a complementary log-log link with the log of interval length included as an offset. Predictor variables included species identity, incubation initiation date, daily incubation constancy (logit-transformed), and daily recess frequency. Day of year was also evaluated as a seasonal covariate. Continuous predictors were standardized prior to analysis to facilitate comparison of effect sizes and improve numerical stability. We compared a set of iterative candidate models representing alternative hypotheses about drivers of nest failure. Model selection was based on AICc and inference focused on relative support among models. To account for within-nest non-independence among exposure intervals, we calculated cluster-robust standard errors at the nest level, ensuring valid inference despite repeated observations within nests.

```{r}
# Create boundary adjustment function
eps <- 0.001

logit_eps <- function(p, eps = 0.001) {
  p_adj <- pmin(pmax(p, eps), 1 - eps)
  log(p_adj / (1 - p_adj))
}

# Nest-level fate table
nest_fate <- hen.act.filtered %>%
  mutate(
    DateTimeBegin = suppressWarnings(lubridate::ymd_hms(DateTimeBegin, tz = "UTC")),
    date = as.Date(DateTimeBegin)
  ) %>%
  group_by(Nest_ID) %>%
  summarise(
    Species        = first(Species),
    Fate_simple    = first(na.omit(Fate_simple)),
    inc_init_final = first(na.omit(inc_init_final)),
    fate_date      = suppressWarnings(max(date, na.rm = TRUE)),
    n_dates        = sum(!is.na(date)),
    .groups = "drop"
  ) %>%
  filter(
    Fate_simple %in% c("Hatched", "Failed"),
    n_dates > 0,
    is.finite(fate_date),
    !is.na(inc_init_final)
  ) %>%
  mutate(
    Species = factor(Species),
    Nest_ID = factor(Nest_ID)
  )

# Join nest fate to daily predictors
le0 <- daily_wx %>%
  transmute(
    Species = factor(Species),
    Nest_ID = factor(Nest_ID),
    date    = as.Date(date),
    DOY     = as.integer(DOY),
    constancy_daily = as.numeric(constancy_daily),
    recesses_daily  = as.numeric(recesses_daily)
  ) %>%
  mutate(
    constancy_daily_logit = if_else(
      is.na(constancy_daily),
      NA_real_,
      logit_eps(constancy_daily, eps)
    )
  ) %>%
  left_join(
    nest_fate %>% select(Nest_ID, Fate_simple, fate_date, inc_init_final),
    by = "Nest_ID"
  ) %>%
  filter(
    !is.na(Fate_simple),
    !is.na(fate_date),
    date <= fate_date
  ) %>%
  arrange(Nest_ID, date)

# Convert to exposure intervals
le_dat <- le0 %>%
  group_by(Nest_ID) %>%
  mutate(
    Year = factor(lubridate::year(date)),
    date_next     = lead(date),
    exposure_days = as.numeric(difftime(date_next, date, units = "days")),
    keep_interval = !is.na(date_next) &
                    is.finite(exposure_days) &
                    exposure_days > 0
  ) %>%
  ungroup() %>%
  filter(keep_interval) %>%
  group_by(Nest_ID) %>%
  mutate(
    is_last_interval = row_number() == n()
  ) %>%
  ungroup() %>%
  mutate(
    alive_next   = if_else(Fate_simple == "Failed" & is_last_interval, 0L, 1L),
    log_exposure = log(exposure_days),
    z_constancy  = as.numeric(scale(constancy_daily_logit)),
    z_DOY        = as.numeric(scale(DOY)),
    z_recesses   = as.numeric(scale(recesses_daily)),
    z_inc_init   = as.numeric(scale(inc_init_final))
  ) %>%
  filter(
    is.finite(log_exposure),
    !is.na(z_constancy),
    !is.na(z_DOY),
    !is.na(z_recesses),
    !is.na(z_inc_init)
  )

# Manual AICc function
AICc_manual <- function(logLik, K, n) {
  aic <- -2 * logLik + 2 * K
  if ((n - K - 1) <= 0) return(NA_real_)
  aic + (2 * K * (K + 1)) / (n - K - 1)
}

# Logistic-exposure GLMs with cloglog link
glm_forms <- list(
  null              = alive_next ~ 1 + offset(log_exposure),
  species           = alive_next ~ Species + offset(log_exposure),
  doy_species       = alive_next ~ z_DOY + Species + offset(log_exposure),
  const_species     = alive_next ~ z_constancy + Species + offset(log_exposure),
  recess_species    = alive_next ~ z_recesses + Species + offset(log_exposure),
  incinit_species   = alive_next ~ z_inc_init + Species + offset(log_exposure),
  doy_const_species = alive_next ~ z_DOY + z_constancy + Species + offset(log_exposure)
)

safe_glm <- function(formula) {
  tryCatch(
    glm(formula,
        family = binomial(link = "cloglog"),
        data = le_dat),
    error = function(e) NULL
  )
}

glm_mods <- purrr::imap(glm_forms, ~ safe_glm(.x))


# Model selection table
glm_tab <- purrr::imap_dfr(glm_mods, function(m, nm) {
  if (is.null(m)) {
    return(tibble(model = nm, ok = FALSE))
  }
  ll <- as.numeric(logLik(m))
  K  <- length(coef(m))
  n  <- nobs(m)
  tibble(
    model  = nm,
    ok     = TRUE,
    n      = n,
    K      = K,
    logLik = ll,
    AIC    = AIC(m),
    AICc   = AICc_manual(ll, K, n)
  )
})

use_AICc <- any(is.finite(glm_tab$AICc))

glm_tab <- glm_tab %>%
  mutate(criterion = if (use_AICc) AICc else AIC) %>%
  arrange(criterion) %>%
  mutate(
    delta  = criterion - min(criterion, na.rm = TRUE),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta), na.rm = TRUE)
  )

# publication-quality model results table
model_labels <- c(
  null              = "Intercept",
  species           = "Intercept + Species",
  doy_species       = "Intercept + DOY + Species",
  const_species     = "Intercept + Constancy + Species",
  recess_species    = "Intercept + Recesses + Species",
  incinit_species   = "Intercept + Initiation + Species",
  doy_const_species = "Intercept + DOY + Constancy + Species"
)

tab_le_aic <- glm_tab %>%
  filter(ok) %>%
  mutate(
    Model = dplyr::recode(model, !!!model_labels),
    criterion = if (use_AICc) AICc else AIC
  ) %>%
  arrange(criterion) %>%
  mutate(
    `ΔAICc` = criterion - min(criterion, na.rm = TRUE),
    Weight  = exp(-0.5 * `ΔAICc`) / sum(exp(-0.5 * `ΔAICc`), na.rm = TRUE)
  )

ic_col <- if (use_AICc) "AICc" else "AIC"

Table5 <- tab_le_aic %>%
  transmute(
    Model,
    K,
    logLik,
    !!ic_col := .data[[ic_col]],
    `ΔAICc`,
    Weight
  ) %>%
  gt() %>%
  cols_align(align = "center", columns = everything()) %>%
  cols_align(align = "left", columns = Model) %>%
  fmt_number(columns = "K", decimals = 0) %>%
  fmt_number(columns = "logLik", decimals = 2) %>%
  fmt_number(columns = c(ic_col, "ΔAICc"), decimals = 2) %>%
  fmt_number(columns = "Weight", decimals = 3) %>%
  cols_label(
    logLik = html("logLik"),
    Weight = html("w<sub>i</sub>"),
    `ΔAICc` = html("&Delta;AICc")
  ) %>%
  tab_options(
    table.font.size  = px(12),
    data_row.padding = px(2)
  )

Table5

# export table for manuscript
gtsave(
  Table5,
  filename = "Table5.png",
  vwidth  = 900,
  vheight = 300
)
```

**Table 5.** Model selection results for logistic-exposure models of daily nest survival, ranked by AICc for Arctic-breeding sea ducks in Utqiagvik, Alaska, USA between 2011-2025.

```{r}
# supporting figure
# Predicted daily nest survival vs incubation initiation. 
# pick best model
m_best <- glm_mods[["incinit_species"]]

# inverse cloglog -> daily survival
inv_cloglog <- function(eta) 1 - exp(-exp(eta))

# model Species levels (critical: must match fit)
species_levels <- m_best$xlevels$Species
if (is.null(species_levels)) species_levels <- levels(m_best$model$Species)

# build an initiation grid on the ORIGINAL scale, then convert to z using le_dat params
inc_mu <- mean(le_dat$inc_init_final, na.rm = TRUE)
inc_sd <- sd(le_dat$inc_init_final, na.rm = TRUE)

inc_grid <- seq(
  quantile(le_dat$inc_init_final, 0.02, na.rm = TRUE),
  quantile(le_dat$inc_init_final, 0.98, na.rm = TRUE),
  length.out = 160
)

# exposure = 1 day => log_exposure = 0
newdat <- expand_grid(
  Species = factor(species_levels, levels = species_levels),
  inc_init_final = inc_grid
) %>%
  mutate(
    z_inc_init   = (inc_init_final - inc_mu) / inc_sd,
    log_exposure = 0
  )

# predict on link scale, then transform
pr <- predict(m_best, newdata = newdat, type = "link", se.fit = TRUE)

fig_dat <- newdat %>%
  mutate(
    eta    = pr$fit,
    se     = pr$se.fit,
    lo_eta = eta - 1.96 * se,
    hi_eta = eta + 1.96 * se,
    S_day  = inv_cloglog(eta),
    lo     = inv_cloglog(lo_eta),
    hi     = inv_cloglog(hi_eta)
  )

# plot
fig_init <- ggplot(fig_dat, aes(x = inc_init_final, y = S_day,
                                color = Species, fill = Species)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = palette, labels = species_labels) +
  scale_fill_manual(values = palette, labels = species_labels) +
  scale_y_continuous(limits = c(0.4, 1),
                     labels = percent_format(accuracy = 1)) +
  labs(
    x = "Incubation initiation date",
    y = "Predicted daily nest survival",
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(face = "bold")
)

fig_init
ggsave(
  "Fig6.png",
  plot = fig_init,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 6.** Predicted daily nest survival as a function of incubation initiation date for four Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. Lines show species-specific model predictions, illustrating declining daily nest survival with later incubation initiation. On the x-axis, days range from 160 (June 9) and 190 (July 9).

**Best-fit model diagnostics**

```{r}
# choose best supported model, w/ name
glm_tab_clean <- glm_tab %>% filter(ok, is.finite(AICc)) %>% arrange(AICc)
best_name <- glm_tab_clean$model[1]
m_best_le <- glm_mods[[best_name]]

best_name
summary(m_best_le)

summary(le_dat$exposure_days)
quantile(le_dat$exposure_days, probs = c(0, .05, .25, .5, .75, .95, 1), na.rm = TRUE)

# Event rate by species
le_dat %>%
  group_by(Species) %>%
  summarise(
    n_int = n(),
    events = sum(alive_next == 0),
    event_rate = mean(alive_next == 0),
    .groups = "drop"
  )

# convergence / numerical stability
summary(m_best_le)$coefficients

# Check for very large coefficients / SEs
coef_table <- broom::tidy(m_best_le)
coef_table %>% arrange(desc(abs(estimate))) %>% head(10)
coef_table %>% arrange(desc(std.error)) %>% head(10)

# Check dispersion/overdispersion due to clustering (intervals within nests)
performance::check_overdispersion(m_best_le)

# compute a simple dispersion ratio (Pearson chi-square / df)
pearson_dispersion <- function(m) {
  rp <- residuals(m, type = "pearson")
  chisq <- sum(rp^2)
  df <- df.residual(m)
  tibble(chisq = chisq, df = df, dispersion = chisq / df)
}
pearson_dispersion(m_best_le)
# If dispersion > 1.2, consider cluster-robust SEs

# simulation-based residual diagnostics
set.seed(1)
sim <- DHARMa::simulateResiduals(fittedModel = m_best_le, n = 1000)
plot(sim)

DHARMa::testDispersion(sim)
DHARMa::testOutliers(sim)
DHARMa::testUniformity(sim)

# Collinearity among predictors
car::vif(m_best_le)

# check simple correlations among numeric covariates you use together:
le_dat %>%
  select(z_DOY, z_constancy, z_recesses, z_inc_init) %>%
  cor(use = "pairwise.complete.obs")

# Check clustering / non-independence by Nest_ID
# Because intervals are nested within Nest_ID, residuals may be correlated.
# average DHARMa residuals by nest should be centered ~0.5.
res_by_nest <- tibble(
  Nest_ID = le_dat$Nest_ID,
  simres  = sim$scaledResiduals
) %>%
  group_by(Nest_ID) %>%
  summarise(mean_resid = mean(simres), n_int = n(), .groups = "drop")

summary(res_by_nest$mean_resid)

ggplot(res_by_nest, aes(x = n_int, y = mean_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = 2) +
  labs(x = "Intervals per nest", y = "Mean DHARMa residual by nest") +
  theme_minimal(base_size = 12)

```

Model diagnostics indicated adequate fit and no substantial overdispersion, apart from some clustering. Because exposure intervals were nested within nests, we proceed to account for within-nest non-independence using cluster-robust standard errors at the nest level (Shaffer et al. 2004).

```{r}
# Effect, coefficient, standard error table 

m <- m_best_le

# model pieces
X <- model.matrix(m)
u <- residuals(m, type = "response")
mu <- fitted(m)
nest <- le_dat$Nest_ID

stopifnot(nrow(X) == length(u), nrow(X) == length(nest))

# (X' W X)^-1
W <- as.numeric(mu * (1 - mu))
XtW <- t(X) * W
bread <- solve(XtW %*% X)

# Clustered score outer products
idx_by_cluster <- split(seq_len(nrow(X)), nest)

S <- vapply(
  idx_by_cluster,
  function(idx) colSums(X[idx, , drop = FALSE] * u[idx]),
  FUN.VALUE = numeric(ncol(X))
)

meat <- S %*% t(S)

# CR1 small-sample correction
G <- length(idx_by_cluster)
N <- nrow(X)
p <- ncol(X)
dfc <- (G / (G - 1)) * ((N - 1) / (N - p))

V_cr <- dfc * (bread %*% meat %*% bread)

# Coefficients  and robust SEs
b <- coef(m)

se <- sqrt(diag(V_cr))
names(se) <- colnames(X)
se <- se[names(b)]

tab <- tibble(
  term = names(b),
  coef = as.numeric(b),
  se   = as.numeric(se)
)

# Label effects via named-vector lookup
effect_labels <- c(
  "(Intercept)" = "Intercept",
  "z_inc_init"  = "Incubation initiation date",
  "SpeciesLTDU" = "  Long-tailed Duck",
  "SpeciesSPEI" = "  Spectacled Eider",
  "SpeciesSTEI" = "  Steller's Eider"
)

tab <- tab %>%
  mutate(
    Effect = ifelse(term %in% names(effect_labels),
                    effect_labels[term],
                    term)
  ) %>%
  select(Effect, coef, se)

# table formatting

# row order
tab <- tab %>%
  mutate(
    Effect = factor(
      Effect,
      levels = c(
        "Intercept",
        "  Long-tailed Duck",
        "  Spectacled Eider",
        "  Steller's Eider",
        "Incubation initiation date"
      )
    )
  ) %>%
  arrange(Effect) %>%
  mutate(Effect = as.character(Effect))

# build pub-quality table

table_input <- tab %>%
  transmute(
    Effect = Effect,
    Coefficient = coef,
    `Standard error` = se
  ) %>%
  filter(!is.na(Effect))

Table_coefficients <- gt::gt(table_input) %>%
  gt::fmt_number(columns = c("Coefficient", "Standard error"), decimals = 3) %>%
  gt::fmt_missing(columns = c("Coefficient", "Standard error"), missing_text = "") %>%
  gt::cols_align(align = "left",   columns = "Effect") %>%
  gt::cols_align(align = "center", columns = c("Coefficient", "Standard error")) %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(rows = Effect %in% c("Intercept", "Incubation initiation date"))
  ) %>%
  gt::tab_options(
    table.font.size = gt::px(12),
    data_row.padding = gt::px(2),
    table.border.top.style = "solid",
    table.border.bottom.style = "solid",
    column_labels.border.bottom.style = "solid",
    table_body.hlines.style = "none"
  )

Table_coefficients

# export table for manuscript
gtsave(
  Table_coefficients,
  filename = "Table6.png",
  vwidth  = 900,
  vheight = 300
)
```

**Table 6.** Coefficient estimates and standard errors from logistic exposure model relating daily nest survival to incubation initiation date for Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. The intercept represents *S. spectabilis*, with coefficients for other species expressed as deviations from this reference category.

**Key Model Takeaways**

-   Temporal drivers (initiation dates) and behavioral drivers (recesses) from top 2 models explain similar variation in daily nest survival, once species differences are accounted for.

-   Support for strong interspecific differences in baseline daily nest survival (species present in all best candidate models).

-   Monotonic declines in predicted daily nest survival with later incubation initiation for all species. Parallel slopes indicating similar seasonal effect but different intercepts. Earlier-nesting females consistently experience higher daily nest survival.

-   For discussion, why might later nesting females experience lower daily nest survival: warmer temps, different predator encounter rates, insect harassment, lower reserves?
