---
title: "Updated code from Chris's"
author: "Kayla Shively"
format: html
editor: visual
---

```{r}
# libraries

library(dplyr)
library(ggplot2)
library(lubridate)
library(gt)
library(purrr)
library(tidyr)
library(lme4)
library(MuMIn)
library(ggeffects)
library(scales)
library(tibble)

# Set color palette
palette <- c(
  "KIEI" = "#4DBBD5FF",  
  "LTDU" = "#3C5488FF",  
  "SPEI" = "#00A087FF", 
  "STEI" = "#E64B35FF"   
)

pal_common <- c(
  "King Eider"         = unname(palette["KIEI"]),
  "Long-tailed Duck"   = unname(palette["LTDU"]),
  "Steller's Eider"    = unname(palette["STEI"]),
  "Spectacled Eider"   = unname(palette["SPEI"]))

# add common names
species_labels <- c(
  KIEI = "King Eider",
  LTDU = "Long-tailed Duck",
  STEI = "Steller's Eider",
  SPEI = "Spectacled Eider"
)

```

## Summaries

```{r}
### Create parsed datetimes and durations, keeping AK time zone
hen_act_sql_time_dur <- hen_act_sql %>%
  mutate(
    begin_dt = with_tz(ymd_hms(datetime_begin_corrected, quiet = TRUE), tzone = "America/Anchorage"),
    end_dt   = with_tz(ymd_hms(datetime_end_corrected, quiet = TRUE),   tzone = "America/Anchorage"),
    duration_min = as.numeric(difftime(end_dt, begin_dt, units = "mins")),
    duration_hr  = duration_min / 60
  )

### Plot incubation bout durations
hen_act_sql_time_dur %>%
  filter(behavior == "I-Incubating") %>%
  ggplot(aes(duration_hr)) +
  geom_histogram(bins = 100) +
  coord_cartesian(xlim = c(0, 150)) +
  labs(
    x = "Incubation bout length (hours)",
    y = "Frequency",
    title = "Distribution of incubation bout durations"
  ) +
  theme_minimal()

### Check for absurb values
hen_act_sql_time_dur %>%
  filter(behavior == "I-Incubating") %>%
  summarise(
    min_hr = min(duration_hr, na.rm = TRUE),
    mean_hr = mean(duration_hr, na.rm = TRUE),
    max_hr = max(duration_hr, na.rm = TRUE),
    n_over_24 = sum(duration_hr > 24, na.rm = TRUE)
  )

behaviors_over_24hr <- hen_act_sql_time_dur %>%
  filter(behavior == "I-Incubating",
    !is.na(duration_hr),
    duration_hr > 24
  ) %>%
  arrange(desc(duration_hr))
```

```{r}
### Identify nests with <1440 minutes of observation
View(hen_act_sql_time_dur)

hen_act_sql_time_dur_total <- hen_act_sql_time_dur %>%
  group_by(nest_id) %>%
  mutate(
    total_mins = sum(
      duration_min[behavior %in% c("N-NotIncubating", "I-Incubating")],
      na.rm = TRUE
    )
  ) %>%
  ungroup()

## Identify valid nests (>= 24 hours of observation)
valid_nests <- hen_act_sql_time_dur_total %>%
  filter(total_mins >= 1440) %>%
  distinct(nest_id) %>%
  pull(nest_id)

## New working dataset without low-observation nests
hen_act_sql_time_dur_total_valid <- hen_act_sql_time_dur_total %>%
  filter(nest_id %in% valid_nests)

View(hen_act_sql_time_dur_total_valid)
```

Filtering

-   `` `r length(unique(hen_act_sql_time_dur_total_valid$nest_id)) - length(unique(hen_act_sql_time_dur_total$nest_id))` `` nests were removed from the dataset because they had \<1 day (1440 minutes) of incubation behavior (failed shortly after deployment).

**Table 1.** Number of sea duck nests monitored with time-lapse cameras from 2011-2025, after filtering nests with less than 1 day of incubation monitoring. Apparent nest success is calculated as the proportion of nests that hatched among those with known fates. Years without data reflect periods of experimental nest covers or suspended field operations (2017, 2018, 2020, and 2021).

```{r}
### Add species and year into activities for summary
hen_nests_joined <- hen_act_sql_time_dur_total_valid %>%
  inner_join(nests_master_clean, by = "nest_id")

## Check number of nests retained
hen_nests_joined %>%
  summarise(n_nests = n_distinct(nest_id))

## Check if any init values are missing
hen_nests_joined %>%
  summarise(missing_init_median = sum(is.na(init_median)))


# Create one record per nest
nest_level <- hen_nests_joined %>%
  group_by(species, year, nest_id) %>%
  summarise(
    fate_simple = case_when(
      any(fate == "Hatched", na.rm = TRUE) ~ "Hatched",
      any(fate == "Destroyed", na.rm = TRUE) ~ "Destroyed",
      TRUE ~ NA_character_
    ),
    .groups = "drop"
  )

# Count nests per species per year
nests_per_year <- nest_level %>%
  count(species, year, name = "n_nests")

# Calculate apparent nest success
apparent_success <- nest_level %>%
  filter(!is.na(fate_simple)) %>%
  group_by(species) %>%
  summarise(
    apparent_success = mean(fate_simple == "Hatched"),
    apparent_success_pct = round(100 * apparent_success, 1),
    .groups = "drop"
  )

# Convert to wide format
nests_wide <- nests_per_year %>%
  pivot_wider(
    names_from = year,
    values_from = n_nests,
    values_fill = 0
  )

# Order year columns numerically
year_cols <- nests_wide %>%
  select(-species) %>%
  names() %>%
  as.integer() %>%
  sort() %>%
  as.character()

nests_wide_ordered <- nests_wide %>%
  select(species, all_of(year_cols))

# Combine tables
summary_wide <- nests_wide_ordered %>%
  left_join(apparent_success, by = "species") %>%
  mutate(
    Total = rowSums(across(all_of(year_cols)), na.rm = TRUE)
  ) %>%
  relocate(Total, .after = all_of(tail(year_cols, 1))) %>%
  arrange(species)

# Convert species codes to common names
nest_summary <- summary_wide %>%
  mutate(
    species = dplyr::recode(
      as.character(species),
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      STEI = "Steller's Eider",
      SPEI = "Spectacled Eider"
    )
  ) %>%
  select(species, all_of(year_cols), Total, apparent_success_pct) %>%
  rename(
    Species = species,
    `Apparent Nest Success (%)` = apparent_success_pct
  ) %>%
  gt() %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(2)
  )

nest_summary

# Save table
gtsave(
  nest_summary,
  filename = "table_nest_summary_by_species_year.png",
  vwidth = 900,
  vheight = 300
)
View(hen_nests_joined)
```

\
**Table 2.** Coverage of time-lapse video monitoring of incubation behavior in Arctic-breeding sea duck nests near Utqiagvik, Alaska, USA between 2011-2025.

```{r}
# If needed, first create duration_days from duration_hr
hen_nests_joined <- hen_nests_joined %>%
  mutate(duration_days = duration_min / 1440)

# Summarize days per nest of I and N behaviors
nest_days <- hen_nests_joined %>%
  filter(behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  group_by(species, nest_id) %>%
  summarise(
    days_per_nest = sum(duration_days, na.rm = TRUE),
    .groups = "drop"
  )

# Summarize by species and add SD
nest_summary <- nest_days %>%
  group_by(species) %>%
  summarise(
    Number_of_Nests = n(),
    Total_Days_Monitored = sum(days_per_nest, na.rm = TRUE),
    Mean_Days = mean(days_per_nest, na.rm = TRUE),
    SD_Days = sd(days_per_nest, na.rm = TRUE),
    Mean_SD = paste0(
      round(Mean_Days, 1), " ± ", round(SD_Days, 1)
    ),
    .groups = "drop"
  )

# Create table with mean ± SD instead of separate columns
nest_coverage <- nest_summary %>%
  mutate(
    species = dplyr::recode(
      as.character(species),
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      STEI = "Steller's Eider",
      SPEI = "Spectacled Eider"
    )
  ) %>%
  rename(Species = species) %>%
  select(
    Species,
    `N (nests)` = Number_of_Nests,
    `Days Monitored` = Total_Days_Monitored,
    `Mean (± SD)` = Mean_SD
  ) %>%
  gt() %>%
  cols_align(align = "center", columns = everything()) %>%
  fmt_number(columns = c(`N (nests)`, `Days Monitored`), decimals = 0) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(2)
  )

nest_coverage

gtsave(
  nest_coverage,
  filename = "table_nest_coverage.png",
  vwidth = 900,
  vheight = 300
)
```

```{r}
# Calculate days since initiation
hen_nests_joined <- hen_nests_joined %>%
  mutate(
    begin_dt = if ("begin_dt" %in% names(.)) begin_dt else ymd_hms(datetime_begin_corrected, quiet = TRUE),
    doy = yday(begin_dt),
    day_since_init = doy - init_median,
    day_since_init_int = floor(day_since_init)
  )

# Build dataset for nests monitored per day since initiation
monitoring_dist <- hen_nests_joined %>%
  filter(
    behavior %in% c("I-Incubating", "N-NotIncubating"),
    !is.na(day_since_init_int),
    !is.na(species),
    !is.na(nest_id)
  ) %>%
  distinct(species, nest_id, day_since_init_int) %>%   # one record per nest per day
  count(species, day_since_init_int, name = "n_nests")

# Clean species labels for plotting
plot_dat <- monitoring_dist %>%
  mutate(
    species = trimws(as.character(species)),
    Species = case_when(
      species == "KIEI" ~ "King Eider",
      species == "LTDU" ~ "Long-tailed Duck",
      species == "STEI" ~ "Steller's Eider",
      species == "SPEI" ~ "Spectacled Eider",
      TRUE ~ NA_character_
    ),
    Species = factor(
      Species,
      levels = c("King Eider", "Long-tailed Duck", "Steller's Eider", "Spectacled Eider")
    )
  ) %>%
  filter(!is.na(Species))

nest_coverage <- ggplot(plot_dat, aes(x = day_since_init_int, y = n_nests, fill = Species)) +
  geom_col() +
  facet_wrap(~ Species, ncol = 2, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "King Eider"       = unname(palette["KIEI"]),
      "Long-tailed Duck" = unname(palette["LTDU"]),
      "Steller's Eider"  = unname(palette["STEI"]),
      "Spectacled Eider" = unname(palette["SPEI"])
    ),
    drop = FALSE
  ) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, NA)) +
  labs(
    x = "Days since nest initiation",
    y = "Number of nests monitored",
    fill = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )

nest_coverage

ggsave(
  "nest_coverage.png",
  plot = nest_coverage,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 3.** Distribution of monitoring duration across the incubation period for four Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. Histograms show the number of nests monitored by day since nest initiation for *S. spectabilis*, *C. hyemalis*, *P. stelleri,* and *S. fischeri*, illustrating camera monitoring coverage and temporal distribution during incubation.

**Table 3.** Mean values of incubation constancy of Hatched nests with 95% confidence intervals for Arctic-breeding sea duck nests near Utqiagvik, Alaska, USA between 2011-2025. Event-weighted, not nest-weighted.

```{r}
####### You may need to address unviewable nest days prior to these analyses. 

### Calculate constancy
constancy_events <- hen_nests_joined %>%
  filter(
    fate == "Hatched",
    behavior %in% c("I-Incubating", "N-NotIncubating"),
    !is.na(duration_min),
    duration_min >= 0
  ) %>%
  mutate(
    incubating_time = if_else(behavior == "I-Incubating", duration_min, 0),
    recess_time     = if_else(behavior == "N-NotIncubating", duration_min, 0)
  )

constancy_summary <- constancy_events %>%
  group_by(species) %>%
  summarise(
    incubating_minutes = sum(incubating_time, na.rm = TRUE),
    recess_minutes     = sum(recess_time, na.rm = TRUE),
    total_minutes      = incubating_minutes + recess_minutes,
    
    mean_constancy = incubating_minutes / total_minutes,
    
    se_constancy = sqrt(mean_constancy * (1 - mean_constancy) / total_minutes),
    
    CI_lower = mean_constancy - 1.96 * se_constancy,
    CI_upper = mean_constancy + 1.96 * se_constancy,
    
    .groups = "drop"
  )

constancy_summary <- constancy_summary %>%
  mutate(
    Species = dplyr::recode(
      as.character(species),
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      STEI = "Steller's Eider",
      SPEI = "Spectacled Eider"
    )
  )

constancy_table <- constancy_summary %>%
  select(
    Species,
    `Mean constancy` = mean_constancy,
    `Lower 95% CI` = CI_lower,
    `Upper 95% CI` = CI_upper
  ) %>%
  gt() %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  fmt_number(
    columns = c(`Mean constancy`, `Lower 95% CI`, `Upper 95% CI`),
    decimals = 3
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(2)
  )

constancy_table

gtsave(
  constancy_table,
  filename = "constancy_table.png",
  vwidth = 900,
  vheight = 300
)
```

```{r}
### Comparing Hatched and Destroyed Nests
# Nest-level incubation constancy
nest_behavior_summary <- hen_nests_joined %>%
  filter(
    behavior %in% c("I-Incubating", "N-NotIncubating"),
    fate %in% c("Hatched", "Destroyed"),
    !is.na(duration_min),
    duration_min >= 0
  ) %>%
  group_by(species, nest_id, fate, behavior) %>%
  summarise(
    total_minutes = sum(duration_min, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = behavior,
    values_from = total_minutes,
    values_fill = 0
  ) %>%
  mutate(
    total_monitored = `I-Incubating` + `N-NotIncubating`,
    constancy = if_else(
      total_monitored > 0,
      `I-Incubating` / total_monitored,
      NA_real_
    )
  ) %>%
  filter(!is.na(constancy))

# Extract nest-level constancy plus fate
nest_df <- nest_behavior_summary %>%
  select(species, nest_id, fate, constancy)

# Prep plot dataframe
df <- nest_df %>%
  mutate(
    Species = dplyr::recode(
      as.character(species),
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      STEI = "Steller's Eider",
      SPEI = "Spectacled Eider"
    ),
    Species = factor(
      Species,
      levels = c("King Eider", "Long-tailed Duck", "Steller's Eider", "Spectacled Eider")
    ),
    Fate_simple = factor(fate, levels = c("Destroyed", "Hatched"))
  ) %>%
  filter(!is.na(constancy), !is.na(Species), !is.na(Fate_simple))

# Group summaries (mean ± 95% CI)
sum_sf <- df %>%
  group_by(Species, Fate_simple) %>%
  summarise(
    n = n(),
    mean = mean(constancy, na.rm = TRUE),
    sd = sd(constancy, na.rm = TRUE),
    se = sd / sqrt(n),
    tcrit = qt(0.975, df = pmax(n - 1, 1)),
    lo = mean - tcrit * se,
    hi = mean + tcrit * se,
    .groups = "drop"
  )

# Wilcoxon within species and labels
p_lab <- df %>%
  group_by(Species) %>%
  summarise(
    p_value = if_else(
      all(c("Destroyed", "Hatched") %in% Fate_simple) &
        sum(Fate_simple == "Destroyed") > 0 &
        sum(Fate_simple == "Hatched") > 0,
      wilcox.test(constancy ~ Fate_simple, exact = FALSE)$p.value,
      NA_real_
    ),
    y_pos = max(constancy, na.rm = TRUE) + 0.05 * diff(range(constancy, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    label = case_when(
      is.na(p_value)  ~ "p = NA",
      p_value < 0.001 ~ "p < 0.001",
      TRUE ~ paste0("p = ", signif(p_value, 2))
    )
  ) %>%
  mutate(y_pos = pmin(y_pos, 0.98))

# Set palette
pal_common <- c(
  "King Eider" = unname(palette["KIEI"]),
  "Long-tailed Duck" = unname(palette["LTDU"]),
  "Steller's Eider" = unname(palette["STEI"]),
  "Spectacled Eider" = unname(palette["SPEI"])
)

# Plot figure
Fig3 <- ggplot(df, aes(x = Fate_simple, y = constancy, color = Species)) +
  geom_errorbar(
    data = sum_sf,
    aes(x = Fate_simple, ymin = lo, ymax = hi, color = Species),
    width = 0.18,
    linewidth = 1.1,
    inherit.aes = FALSE
  ) +
  geom_point(
    position = position_jitter(width = 0.08, height = 0),
    alpha = 0.6,
    size = 2
  ) +
  geom_text(
    data = p_lab,
    aes(x = 1.5, y = y_pos, label = label),
    inherit.aes = FALSE,
    size = 4,
    color = "grey20"
  ) +
  facet_wrap(~ Species, ncol = 2) +
  scale_color_manual(values = pal_common, guide = "none") +
  coord_cartesian(ylim = c(0.60, 1.00)) +
  labs(
    x = "Nest fate",
    y = "Incubation constancy"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )

Fig3

# Export as PNG
ggsave(
  "Fig3.png",
  plot = Fig3,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 3.** Mean daily incubation constancy by nest fate for four Arctic-breeding sea ducks near Utqiaġvik, Alaska between 2011-2025. Points represent nest-level mean incubation constancy (excluding the terminal day), with larger points and error bars indicating species-specific means ± SD for failed and hatched nests. P-values denote results of species-specific comparisons between nest fate categories.

```{r}
core_frac <- 0.50  # dominant window contains this fraction of recesses

# one row per recess with hour of day (common names)
recess_tod <- hen_nests_joined %>%
  filter(behavior %in% c("N-NotIncubating", "not_incubating")) %>%
  mutate(
    begin_dt = if ("begin_dt" %in% names(.)) {
      begin_dt
    } else {
      ymd_hms(datetime_begin_corrected, quiet = TRUE)
    },
    hour = lubridate::hour(begin_dt),
    Species = dplyr::recode(
      as.character(species),
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      STEI = "Steller's Eider",
      SPEI = "Spectacled Eider"
    ),
    Species = factor(
      Species,
      levels = c("King Eider", "Long-tailed Duck", "Steller's Eider", "Spectacled Eider")
    )
  ) %>%
  filter(!is.na(hour), !is.na(Species)) %>%
  distinct(nest_id, Species, activity_id, hour)   # one row per recess event

# counts by hour
hour_counts <- recess_tod %>%
  count(Species, hour, name = "n_recesses") %>%
  complete(Species, hour = 0:23, fill = list(n_recesses = 0L))

Fig4_count <- ggplot(hour_counts, aes(x = hour, y = n_recesses, fill = Species)) +
  geom_col(show.legend = FALSE, width = 0.85) +
  facet_wrap(~ Species, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = pal_common, drop = FALSE) +
  scale_x_continuous(breaks = seq(0, 23, 3), minor_breaks = NULL) +
  labs(
    x = "Hour of day (recess start time)",
    y = "Number of recesses",
    title = "Starting hour of recesses (counts)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

Fig4_count

# proportion by hour within species
hour_summ <- hour_counts %>%
  group_by(Species) %>%
  mutate(prop_recesses = n_recesses / sum(n_recesses, na.rm = TRUE)) %>%
  ungroup()

# helper: find smallest contiguous circular window containing core_frac of recesses
find_core_window <- function(hour, prop, frac = 0.5) {
  full <- tibble(hour = 0:23) %>%
    left_join(tibble(hour = hour, prop = prop), by = "hour") %>%
    mutate(prop = tidyr::replace_na(prop, 0)) %>%
    arrange(hour)

  # duplicate for circular wrap
  h2 <- c(full$hour, full$hour + 24)
  p2 <- c(full$prop, full$prop)

  best_len <- Inf
  best_i <- best_j <- NA
  i <- 1
  s <- 0

  for (j in seq_along(p2)) {
    s <- s + p2[j]
    while (s - p2[i] >= frac) {
      s <- s - p2[i]
      i <- i + 1
    }
    if (s >= frac && h2[i] < 24 && h2[j] <= h2[i] + 24) {
      len <- h2[j] - h2[i]
      if (len < best_len) {
        best_len <- len
        best_i <- i
        best_j <- j
      }
    }
  }

  xmin <- h2[best_i]
  xmax <- h2[best_j] + 1

  if (xmax <= 24) {
    tibble(xmin = xmin, xmax = xmax)
  } else {
    bind_rows(
      tibble(xmin = xmin, xmax = 24),
      tibble(xmin = 0, xmax = xmax - 24)
    )
  }
}

# build per-species shading rectangles
shade_df <- hour_summ %>%
  group_by(Species) %>%
  summarise(
    rects = list(find_core_window(hour, prop_recesses, frac = core_frac)),
    .groups = "drop"
  ) %>%
  unnest(rects) %>%
  mutate(ymin = -Inf, ymax = Inf)

# sample sizes
n_recess_labels <- hour_counts %>%
  group_by(Species) %>%
  summarise(
    n_recess = sum(n_recesses),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("n = ", n_recess))

# proportions + dominant window shading
Fig4 <- ggplot(hour_summ, aes(x = hour, y = prop_recesses, fill = Species)) +
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey30",
    alpha = 0.12
  ) +
  geom_col(show.legend = FALSE, width = 0.85) +
  facet_wrap(~ Species, ncol = 2) +
  scale_fill_manual(values = pal_common, drop = FALSE) +
  scale_x_continuous(breaks = seq(0, 23, 3), minor_breaks = NULL) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Hour of day (recess start time)",
    y = "Proportion of recesses"
  ) +
  geom_text(
    data = n_recess_labels,
    aes(x = 23.5, y = 0.125, label = label),
    inherit.aes = FALSE,
    hjust = 1,
    vjust = 1,
    size = 4
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.4),
    strip.text = element_text(face = "bold")
  )

Fig4

ggsave(
  "Fig4.png",
  plot = Fig4,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

**Figure 4.** Distribution of incubation recess start times across the diel cycle for four Arctic-breeding sea ducks nears Utqiaġvik, Alaska between 2011-2025. Bars show the proportion of observed recesses initiating in each hour of the day. Sample sizes (n) indicate the total number of recess per species. Shaded regions denote the smallest contiguous time interval containing 50% of recess starts.

Daily Incubation Constancy analysis

```{r}
# Logit-normal mixed models for daily incubation constancy
# Response: proportional daily incubation constancy (0–1)
# Model:    lmer() on logit-transformed proportion (logit-normal)
# Random:   nest_id and year, if supported by variance estimates

# read in weather
met_raw <- barrow_ghcnd_daily <- read_csv(
  "C:/Users/Kayla Shively/Desktop/barrow_incubation_analysis/barrow_ghcnd_daily.csv"
)

# data prep: build one row per nest-day
daily_ducks <- hen_nests_joined %>%
  filter(behavior %in% c("I-Incubating", "N-NotIncubating")) %>%
  mutate(
    begin_dt = if ("begin_dt" %in% names(.)) begin_dt else ymd_hms(datetime_begin_corrected, quiet = TRUE),
    date = as.Date(begin_dt)
  ) %>%
  group_by(species, nest_id, date) %>%
  summarise(
    mins_inc = sum(duration_min[behavior == "I-Incubating"], na.rm = TRUE),
    mins_non = sum(duration_min[behavior == "N-NotIncubating"], na.rm = TRUE),
    mins_tot = mins_inc + mins_non,
    constancy_daily = if_else(mins_tot > 0, mins_inc / mins_tot, NA_real_),
    recesses_daily = sum(behavior == "N-NotIncubating", na.rm = TRUE),
    day_since_init = mean(day_since_init, na.rm = TRUE),
    .groups = "drop"
  )

# add in daily weather variables (TMAX + wind)
# drop TMIN due to correlation with TMAX
met_daily <- met_raw %>%
  mutate(
    date = as.Date(DATE),
    TMAX_C = ifelse(is.na(TMAX), NA_real_, TMAX / 10),
    AWND_ms = ifelse(is.na(AWND), NA_real_, AWND / 10)
  ) %>%
  select(date, TMAX_C, AWND_ms) %>%
  arrange(date)

daily_wx <- daily_ducks %>%
  left_join(met_daily, by = "date") %>%
  mutate(
    date = ymd(date),
    DOY = yday(date),
    Year = factor(year(date)),
    Species = factor(species),
    Nest_ID = factor(nest_id)
  )

# standardize analysis dataset and add boundary adjustment for 0/1 values
constancy_camera <- daily_wx %>%
  filter(
    !is.na(constancy_daily),
    !is.na(DOY),
    !is.na(Year),
    !is.na(TMAX_C),
    !is.na(AWND_ms),
    is.finite(day_since_init),
    mins_tot > 0
  ) %>%
  mutate(
    z_DOY = as.numeric(scale(DOY)),
    z_TMAX = as.numeric(scale(TMAX_C)),
    z_AWND = as.numeric(scale(AWND_ms)),
    z_stage = as.numeric(scale(day_since_init)),
    eps = 0.001,
    constancy_adj = pmin(pmax(constancy_daily, eps), 1 - eps),
    logit_const = qlogis(constancy_adj)
  )

# model fitting
ctrl <- lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5)
)

# if Year goes singular, switch to "(1 | Nest_ID)" only
re_form <- "(1 | Nest_ID) + (1 | Year)"

m_cam_null <- lmer(
  as.formula(paste("logit_const ~ 1 +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_stage <- lmer(
  as.formula(paste("logit_const ~ z_stage + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_doy <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_TMAX <- lmer(
  as.formula(paste("logit_const ~ z_stage + z_TMAX + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_AWND <- lmer(
  as.formula(paste("logit_const ~ z_stage + z_AWND + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_TMAX_AWND <- lmer(
  as.formula(paste("logit_const ~ z_stage + z_TMAX + z_AWND + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_doy_TMAX <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + z_TMAX + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_doy_AWND <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + z_AWND + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_doy_TMAX_AWND <- lmer(
  as.formula(paste("logit_const ~ z_DOY + z_stage + z_TMAX + z_AWND + Species +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

m_cam_int_doy <- lmer(
  as.formula(paste("logit_const ~ z_DOY * Species + z_stage + z_TMAX + z_AWND +", re_form)),
  data = constancy_camera,
  control = ctrl,
  REML = FALSE
)

# make AICc table
cam_set <- list(
  null = m_cam_null,
  stage = m_cam_stage,
  doy = m_cam_doy,
  TMAX = m_cam_TMAX,
  wind = m_cam_AWND,
  TMAX_wind = m_cam_TMAX_AWND,
  doy_TMAX = m_cam_doy_TMAX,
  doy_wind = m_cam_doy_AWND,
  doy_TMAX_wind = m_cam_doy_TMAX_AWND,
  int_doy_TMAX_wind = m_cam_int_doy
)

model_labels <- c(
  null = "Intercept (random Nest + Year)",
  stage = "Stage",
  doy = "DOY + stage",
  TMAX = "Stage + TMAX",
  wind = "Stage + wind",
  TMAX_wind = "Stage + TMAX + wind",
  doy_TMAX = "DOY + stage + TMAX",
  doy_wind = "DOY + stage + wind",
  doy_TMAX_wind = "DOY + stage + TMAX + wind",
  int_doy_TMAX_wind = "DOY × Species + stage + TMAX + wind"
)

get_nobs <- function(m) nobs(m)

aicc_cam_tbl <- purrr::imap_dfr(cam_set, \(m, nm) {
  ll <- logLik(m)
  tibble(
    Model_key = nm,
    n = get_nobs(m),
    K = attr(ll, "df"),
    logLik = as.numeric(ll),
    AIC = AIC(m),
    AICc = MuMIn::AICc(m)
  )
}) %>%
  mutate(Model = dplyr::recode(Model_key, !!!model_labels)) %>%
  arrange(AICc) %>%
  mutate(
    `ΔAICc` = AICc - min(AICc, na.rm = TRUE),
    Weight = exp(-0.5 * `ΔAICc`) / sum(exp(-0.5 * `ΔAICc`), na.rm = TRUE),
    Weight = round(Weight, 3)
  )

best_key <- aicc_cam_tbl$Model_key[1]
m_best <- cam_set[[best_key]]

# publication-quality table
Table4 <- aicc_cam_tbl %>%
  select(Model, K, logLik, AICc, `ΔAICc`, Weight) %>%
  gt() %>%
  cols_align(align = "center", columns = c(K, logLik, AICc, `ΔAICc`, Weight)) %>%
  cols_align(align = "left", columns = Model) %>%
  fmt_number(columns = "K", decimals = 0) %>%
  fmt_number(columns = c("logLik", "AICc", `ΔAICc`), decimals = 2) %>%
  fmt_number(columns = "Weight", decimals = 3) %>%
  cols_label(
    Model = "Model",
    K = "K",
    logLik = html("logLik"),
    AICc = html("AIC<sub>c</sub>"),
    `ΔAICc` = html("&Delta;AIC<sub>c</sub>"),
    Weight = html("w<sub>i</sub>")
  ) %>%
  tab_options(
    table.font.size = px(12),
    data_row.padding = px(2)
  )

Table4

gtsave(
  Table4,
  filename = "Table4.png",
  vwidth = 900,
  vheight = 300
)
```

Results :

-   Incubation likely changes across incubation stage, which is biologically expected.

-   DOY x Species interaction

    -   Earlier breeders experience colder temperatures

    -   different species respond differently to weather

    -   timing of snowmelt and food availability differs

-   Weather (TMAX + wind). Aligns with known energetics in KIEI and SPEI.

    -   colder conditions –\> hens may recess less

    -   high wind –\> may increase energetic cost of incubation

Am I overparameterizing? Probably not? AICc penalizes model complexity. My best model has K = 14 parameters, but AICc already penalizes extra parameters. Despite that penalty, the model still wins by \~10 AICc units, which is strong evidence that the added terms improve fit.

```{r}
### MOdel checks

# Is the Year random effect singular? 
isSingular(m_best) # Looking for FALSE

# Look at effect sizes
summary(m_best)

# Check outlier
which.min(residuals(m_best))

isSingular(m_best) # Looking for FALSE
```

Looking at effect sizes:

-   z_DOY:SpeciesLTDU = +0.86

-   z_DOY:SpeciesSTEI = +0.71

-   z_DOY:SpeciesSPEI ≈ 0

KIEI constantly declines through the season, LTDU and STEI have less seasonal declines (possibly stable or increasing), Spectacled has little seasonal change.

-   z_stage = −0.24\
    t ≈ −2.1

Constancy slightly decreases as incubation progresses. Explanations: hens take slightly longer recesses later in incubation, energy depletion, increased embryo thermoregulation near hatch.

Manuscript interpretation: Daily incubation constancy varied among species and across the breeding season. King Eiders exhibited the highest constancy overall, whereas Long-tailed Ducks and Steller's Eiders showed substantially lower constancy. Seasonal patterns differed among species, with constancy declining through the season in King Eiders but remaining relatively stable in other species. Incubation constancy also declined slightly with incubation stage, while daily temperature and wind speed had little effect.

```{r}
### Visualization
# inverse-logit helper
inv_logit <- function(x) plogis(x)

# sequence across observed DOY range
doy_seq <- seq(
  min(constancy_camera$DOY, na.rm = TRUE),
  max(constancy_camera$DOY, na.rm = TRUE),
  length.out = 200
)

# prediction grid
pred_grid <- expand.grid(
  DOY = doy_seq,
  Species = levels(constancy_camera$Species)
) %>%
  as_tibble() %>%
  mutate(
    # recreate z_DOY using the same centering/scaling as the analysis data
    z_DOY = (DOY - mean(constancy_camera$DOY, na.rm = TRUE)) /
      sd(constancy_camera$DOY, na.rm = TRUE),

    # hold other covariates at mean standardized values
    z_stage = 0,
    z_TMAX  = 0,
    z_AWND  = 0
  )

# build design matrix for fixed effects only
X <- model.matrix(
  ~ z_DOY * Species + z_stage + z_TMAX + z_AWND,
  data = pred_grid
)

b <- lme4::fixef(m_best)
V <- as.matrix(vcov(m_best))

# align columns with fixed effects
X <- X[, names(b), drop = FALSE]

# predictions and Wald SE on link scale
fit_link <- as.vector(X %*% b)
se_link  <- sqrt(pmax(0, diag(X %*% V %*% t(X))))

pred_grid <- pred_grid %>%
  mutate(
    fit_link = fit_link,
    se_link  = se_link,
    lo_link  = fit_link - 1.96 * se_link,
    hi_link  = fit_link + 1.96 * se_link,
    fit = inv_logit(fit_link),
    lo  = inv_logit(lo_link),
    hi  = inv_logit(hi_link)
  )

# relabel species with common names
pred_plot <- pred_grid %>%
  mutate(
    Species = recode_factor(
      Species,
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      SPEI = "Spectacled Eider",
      STEI = "Steller's Eider"
    )
  )

# raw data for x-axis rugs
raw_plot <- constancy_camera %>%
  filter(!is.na(DOY), !is.na(Species)) %>%
  mutate(
    Species = recode_factor(
      Species,
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      SPEI = "Spectacled Eider",
      STEI = "Steller's Eider"
    )
  )

# if not already defined
pal_common <- c(
  "King Eider" = unname(palette["KIEI"]),
  "Long-tailed Duck" = unname(palette["LTDU"]),
  "Spectacled Eider" = unname(palette["SPEI"]),
  "Steller's Eider" = unname(palette["STEI"])
)

# publication-quality figure
Fig5 <- ggplot() +
  geom_ribbon(
    data = pred_plot,
    aes(x = DOY, ymin = lo, ymax = hi, fill = Species),
    alpha = 0.25,
    linewidth = 0
  ) +
  geom_rug(
    data = raw_plot,
    aes(x = DOY, color = Species),
    sides = "b",
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  geom_line(
    data = pred_plot,
    aes(x = DOY, y = fit, color = Species),
    linewidth = 1.4
  ) +
  facet_wrap(~ Species, ncol = 2) +
  scale_color_manual(values = pal_common, guide = "none") +
  scale_fill_manual(values = pal_common, guide = "none") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0.60, 1.00),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  labs(
    x = "Day of year (DOY)",
    y = "Predicted daily incubation constancy"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.4, color = "grey80"),
    panel.spacing      = unit(0.9, "lines"),
    strip.text         = element_text(face = "bold")
  )

Fig5

ggsave(
  "Fig5.png",
  plot = Fig5,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  bg = "white"
)
```

This plots the DOY x Species interaction only, with stage, temperature, and wind held at constant. Because my top model is on the logit scale, I back-transformed with plogis().

```{r}
### Plotting stage
stage_seq <- seq(
  min(constancy_camera$day_since_init, na.rm = TRUE),
  max(constancy_camera$day_since_init, na.rm = TRUE),
  length.out = 200
)

pred_stage <- expand.grid(
  day_since_init = stage_seq,
  Species = levels(constancy_camera$Species)
) %>%
  as_tibble() %>%
  mutate(
    z_stage = (day_since_init - mean(constancy_camera$day_since_init, na.rm = TRUE)) /
              sd(constancy_camera$day_since_init, na.rm = TRUE),

    # hold seasonal and weather variables constant
    z_DOY  = 0,
    z_TMAX = 0,
    z_AWND = 0
  )

X <- model.matrix(
  ~ z_DOY * Species + z_stage + z_TMAX + z_AWND,
  data = pred_stage
)

b <- fixef(m_best)
V <- vcov(m_best)

X <- X[, names(b), drop = FALSE]

fit_link <- as.vector(X %*% b)
se_link  <- sqrt(pmax(0, diag(X %*% V %*% t(X))))

pred_stage <- pred_stage %>%
  mutate(
    fit_link = fit_link,
    lo_link  = fit_link - 1.96 * se_link,
    hi_link  = fit_link + 1.96 * se_link,
    fit = plogis(fit_link),
    lo  = plogis(lo_link),
    hi  = plogis(hi_link)
  )

pred_stage <- pred_stage %>%
  mutate(
    Species = recode_factor(
      Species,
      KIEI = "King Eider",
      LTDU = "Long-tailed Duck",
      SPEI = "Spectacled Eider",
      STEI = "Steller's Eider"
    )
  )

Fig_stage <- ggplot(pred_stage,
  aes(x = day_since_init, y = fit, color = Species, fill = Species)) +

  geom_ribbon(aes(ymin = lo, ymax = hi),
              alpha = 0.25,
              linewidth = 0) +

  geom_line(linewidth = 1.4) +

  facet_wrap(~ Species, ncol = 2) +

  scale_color_manual(values = pal_common, guide = "none") +
  scale_fill_manual(values = pal_common, guide = "none") +

  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.60, 1.00)
  ) +

  labs(
    x = "Days since incubation start",
    y = "Predicted incubation constancy"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

Fig_stage
```
